---
title: "Supplementary material for: A BASiCS workflow for expression variability analysis using scRNA-Seq data"
author: 
  - name: Alan O'Callaghan
    affiliation: 
    - &MRC MRC Human Genetics Unit, Institute of Genetics \& Cancer, 
      University of Edinburgh, Western General Hospital, Crewe Road, Edinburgh, 
      EH4 2XU, UK
    email: "a.b.o'callaghan@sms.ed.ac.uk"
  - name: Nils Eling
    affiliation: 
    - &UZH Department of Quantitative Biomedicine, University of Zurich,
      Winterthurerstrasse 190, CH-8057, Zurich, Switzerland
    - &ETH Institute for Molecular Health Sciences, ETH Zurich,
      Otto-Stern Weg 7, 8093 Zurich, Switzerland
  - name: John C. Marioni
    affiliation: 
    - &EBI European Molecular Biology Laboratory, European Bioinformatics 
      Institute, Wellcome Trust Genome Campus, Hinxton, Cambridge CB10 1SD, UK
    - &CRUK Cancer Research UK Cambridge Institute, University of Cambridge, 
      Li Ka Shing Centre, Cambridge, CB2 0RE, UK
  - name: Catalina A. Vallejos
    affiliation: 
    - *MRC 
    - The Alan Turing Institute, British Library, 96 Euston Road, London, 
      NW1 2DB, UK
    email: catalina.vallejos@igmm.ed.ac.uk
bibliography: Workflow.bib
urlcolor: Orange
output:
  BiocWorkflowTools::f1000_article:
    fig_width: 6
    fig_height: 4
---


```{r setup_knitr, include = FALSE, cache = FALSE}
library("BiocStyle")
library("knitr")
knitr::opts_chunk$set(
  message = FALSE, error = FALSE, warning = FALSE,
  cache = 0, cache.path = "cache_supp/",
  fig.path = "figure_supp/"
)
```

# Methods

## The BASiCS model

The BASiCS package features a hierarchical Bayesian model in which the 
transcript count of gene $i$ of cell $j$ is represented as a random variable
$X_{ij}$. For all technical genes, counts can be modelled as Poisson distributed
with a random effect capturing unexplained technical noise:
$$
\begin{aligned}
  X_{ij}|\mu_i,\nu_j & \sim \text{Poisson}(\nu_j\mu_i)\\
  \nu_j|s_j,\theta & \sim \text{Gamma}\left(\frac{1}{\theta},\frac{1}{s_j\theta}\right)
\end{aligned}
$$
where $\mu_i$ explains the gene's mean expression, $\nu_j$ the technical effect
dominated by the mRNA capture efficiency $s_j$ and the unexplained technical 
noise parameter $\theta$. 

This technical influence effects all biological genes in the same manner. 
In addition to technical variations, cell-to-cell differences of biological 
gene expression can be cause by biological processes as explained above. 
To model an additional random effect that captures biological cell-to-cell 
variability, the random variable $X_{ij}$ is modelled as: 
$$
\begin{aligned}
  X_{ij}|\mu_i,\phi_j,\nu_j,\rho_{ij} & \sim \text{Poisson}(\phi_j\nu_j\mu_i\rho_{ij})\\
  \rho_{ij}|\delta_i & \sim \text{Gamma}(\frac{1}{\delta_i}, \frac{1}{\delta_i})
\end{aligned}
$$
where $\phi_j$ is the cell-specifc size factor and $\rho_{ij}$ the biological
random effect incorporating the over-dispersion hyper-parameter $\delta_i$.

In further analysis, $\mu_i$ represents the mean expression and $\delta_i$
the biological over-dispersion of each gene.

### Testing for changes in mean expression and over-dispersion

Differential mean and differential over-dispersion testing is done by computing
the tail posterior propabilities of the difference in mean expression or 
over-dispersion between two conditions ($A$ and $B$) being larger than an
evidence threshold $\tau_0$ or $\omega_0$ [@Bochkina2007, @Vallejos2016]:
$$
\begin{aligned}
  &\text{P}\left(\log\left(\frac{\mu_i^{(A)}}{\mu_i^{(B)}}\right) > \tau_0 | \text{Data}\right) > \alpha_m\\
  &\text{P}\left(\log\left(\frac{\delta_i^{(A)}}{\delta_i^{(B)}}\right) > \omega_0 | \text{Data}\right) > \alpha_d
\end{aligned}
$$

If the tail posterior probability is larger than a given propability threshold 
$\alpha_m$ or $\alpha_d$, the gene is considered to be differentially expressed
or differentially over-dispersed [@Vallejos2016]. The evidence threshold is
usually fixed *a priori* and the probability threshold is defined to
control the expected false discovery rate (EFDR) to 
(e.g. 10%) [@Newton2004, Vallejos2016].

As described by @Vallejos2016, estimates of the
over-dispersion parameters $\delta_i$ are negatively correlated to mean 
expression $\mu_i$. This indicates that in homogeneous populations of cells,
highly expressed genes tend to be less noisy than lowly expressed genes. 
Differential over-dispersion testing is therefore confounded by mean expression
changes. When assessing changes in over-dispersion $\delta_i$, only genes with
no changes in mean expression are considered (see @Vallejos2016).

### Correcting the mean-variability confounding effect

@Eling2018 extended BASiCS to account for the
confounding effect between mean expression and expression variability. 
For this purpose, we capture the relationship between mean and over-dispersion
parameters by introducing the following joint prior distribution for 
$(\mu_i, \delta_i)'$:
$$
  \mu_i \sim \text{log-Normal}\left(0, s^2_{\mu}\right), \hspace{0.5cm}
  \delta_i | \mu_i \sim \text{log-t}_{\eta}\left( \text{f}(\mu_i), \sigma^2 \right).
$$

The latter is equivalent to the non-linear regression model:
$$
  \log(\delta_i) = \text{f}(\mu_i) + \epsilon_i, \hspace{0.5cm} \epsilon_i \sim{}\text{t}_{\eta}(0,\sigma^2), 
$$

where $\mbox{f}(\mu_i)$ represents the over-dispersion (on the log-scale) that
is predicted by the global trend (across all genes) for a given mean expression 
$\mu_i$. Therefore, $\epsilon_i$ can be interpreted as a gene-specific 
*residual over-dispersion* parameter. Positive values for
$\epsilon_i$ indicate more variation than expected for genes with similar 
expression level. Similarly, negative values of $\epsilon_i$ suggest less 
variation than expected [@Eling2018].

In line with the probabilistic approach described above, we identified
statistically significant differences in residual over-dispersion for those 
genes where the tail posterior probability of observing a large difference 
between $\epsilon_i^A$ and $\epsilon_i^B$ exceeds a certain threshold:
$$
  \text{P}(\mid\epsilon_i^{A} - \epsilon_i^{B}\mid > \psi_0 \mid \text{Data}) > \alpha_R,
$$
where $\psi_0 > 0$ defines the minimum tolerance threshold. 
As a default choice, we assume $\psi_0 = \log_2(1.5) / \log_2(e) \approx 0.41$,
which translates into a 50\% increase in over-dispersion. 
The posterior probability threshold $\alpha_R$ is chosen to control the EFDR
(e.g. 10\%) [@Newton2004]. To support interpretability of the results,
we exclude genes that are not expressed in at least 2 cells per condition from
differential variability testing.



# Pre-processing and quality control steps for the presomitic and somitic mesoderm cells

## Loading necessary libraries

```{r libs}
library("SingleCellExperiment")
library("scater")
library("scran")
library("BASiCS")
library("ggplot2")
library("ggpointdensity")
library("biomaRt")
library("patchwork")
library("viridis")
library("readxl")
## set default theme for plots
theme_set(theme_bw())
```

## Obtaining the data

The full dataset is stored under the accession number E-MTAB-6153 on 
ArrayExpress and can be obtained from [here](https://www.ebi.ac.uk/biostudies/arrayexpress/studies/E-MTAB-6153#). 
This workflows uses the *processed* data which consists of 3 files: 

- `rawCounts.tsv`: matrix of raw expression counts (UMI-based) for all genes
and cells.
- `cellAnnotation.tsv`: cell type labels identified by [@Ibarra-Soria2018].
- `normalisedCounts.tsv`: normalised expression counts obtained by 
[@Ibarra-Soria2018]. This file is **not** used in this workflow. 

*Of note:* The `.zip` file is 65 MB in size while the unzipped, raw counts 
measure 873 MB in size.

The code below downloads these as a compressed `.zip` file in the `downloads`
sub-folder. This is decompressed in order to access the individual `.tsv` files. 
After decompression, the `.zip` can be deleted as it is no longer required. 

```{r download-droplet-data}
options(timeout = 1000)
dir.create("downloads", showWarnings = FALSE)
# Check whether the data was downloaded already
if (!file.exists("downloads/rawCounts.tsv")) {
  website <- "https://www.ebi.ac.uk/"
  folder <- "biostudies/files/E-MTAB-6153/"
  file <- "rawCounts.tsv"
  download.file(
    paste0(website, folder, file),
    destfile = "downloads/rawCounts.tsv"
  )
}

```

We then load the raw read counts. 

```{r load-droplet-data-raw}
raw_counts <- read.delim("downloads/rawCounts.tsv", header = TRUE)
```

Prior to using BASiCS, it is important to perform quality control to remove 
poor quality cells and to control for potential heterogeneous substructure 
within the population of cells under study. In this case, [@Ibarra-Soria2018]
has already characterised population structure using clustering techniques. 
Here, we will use such information in order to define the cell populations to
be use as the input for BASiCS. 

The cluster labels identified in the original publication are then loaded:

```{r cluster-labels-droplet}
# Check whether the data was downloaded already
if (!file.exists("downloads/cellAnnotation.tsv")) {
  website <- "https://www.ebi.ac.uk/"
  folder <- "biostudies/files/E-MTAB-6153/"
  file <- "cellAnnotation.tsv"
  download.file(
    paste0(website, folder, file),
    destfile = "downloads/cellAnnotation.tsv"
  )
}
cluster_labels <- read.table("downloads/cellAnnotation.tsv",
  sep = "\t", header = TRUE, stringsAsFactors = FALSE)
```

The data contains the `r length(unique(cluster_labels$cellType))` cell types
identified by the authors. The number of cells per cluster is displayed below:

```{r table-cluster-labels-droplet}
table(cluster_labels$cellType, useNA = "ifany")
```

Cluster labels are not available for `r sum(is.na(cluster_labels$cellType))` cells. 

Note that the authors also identified some sub-structure within these main clusters:

```{r table-subclusters}
# Extract sub-type cell labels (sample sub-index is removed)
cluster_labels$subCellType <- sub("_.*", "", cluster_labels$cell)
table(cluster_labels$subCellType, cluster_labels$cellType)
```

Finally, we also note that the data also contains information about the three
samples from which the cells were extracted. This information can help us to
visualise potential batch effects.

```{r table-batch}
cluster_labels$sample <- as.factor(cluster_labels$sample)
table(cluster_labels$sample)
```

The authors integrated the data from these three samples using the cellranger
aggr program, which uses downsampling to generate samples with similar depth. 

## Select populations of interest

We selected the somitic 
($n = $ `r sum(cluster_labels$cellType == "somiticMesoderm", na.rm = TRUE)`) and 
pre-somitic mesoderm 
($n = $ `r sum(cluster_labels$cellType == "presomiticMesoderm", na.rm = TRUE)`)
cells. This is an arbitrary choice for illustration purposes only. The steps
of this workflow could be applied to any other pair of cell types. 

```{r droplet-cell-selection}
# Identifies which cells are in the groups of interest
ind_som <- which(
  cluster_labels$cellType == "somiticMesoderm" |
  cluster_labels$cellType == "presomiticMesoderm"
)
# Selects the cells of interest from the raw counts and cluster-label metadata
raw_counts <- raw_counts[, ind_som]
cluster_labels <- cluster_labels[ind_som, ]
```

## Generating SingleCellExperiment object 

For pre-processing and visualisation purposes, we load the data into a 
`SingleCellExperiment` object. The metadata will be stored in the `colData` slot.

```{r droplet-sce}
droplet_sce <- SingleCellExperiment(
  assays = list(counts = as(as.matrix(raw_counts), "dgCMatrix")),
  colData = cluster_labels
)
# Delete the original table of raw expression counts as no longer needed
rm(raw_counts)
```

```{r remove-counts}
ind_expressed <- Matrix::rowMeans(counts(droplet_sce)) > 0.1
droplet_sce <- droplet_sce[ind_expressed, ]
```

## Annotation

Input data was annotated using Ensembl gene identifiers. 
To facilitate interpretation, it is often useful to obtain a mapping from 
Ensembl gene IDs to gene symbols using the BioMart suite
([http://www.biomart.org](http://www.biomart.org)) via the 
Bioconductor package `r Biocpkg("biomaRt")` [@Durinck2009].
This can also be used to obtain gene-pathways mappings and other
metadata (e.g. gene length), useful for performing functional analysis
of gene sets identified in downstream analyses.

```{r obtain-gene-symbols}
# Check if data storage folder exists and create it if needed
if (!dir.exists("rds")) {
  dir.create("rds", showWarnings = FALSE)
}
if (!file.exists("rds/genenames.Rds")) {
  # Initialize mart and dataset
  ensembl <- useEnsembl(
    biomart = "genes",
    version = 104,
    dataset = "mmusculus_gene_ensembl"
  )
  # Select gene ID and gene name
  genenames <- getBM(
    attributes = c("ensembl_gene_id", "external_gene_name", "gene_biotype"),
    mart = ensembl
  )
  rownames(genenames) <- genenames$ensembl_gene_id
  # Store information for future use
  saveRDS(genenames, "rds/genenames.Rds")
}
# Load pre-downloaded gene annotation information
genenames <- readRDS("rds/genenames.Rds")
```

We add this information as `rowData` within the `SingleCellExperiment` 
object created above. 

```{r}
## Merge biomaRt annotation
rowdata <- data.frame(ensembl_gene_id = rownames(droplet_sce))
rowdata <- merge(rowdata, genenames, by = "ensembl_gene_id", all.x = TRUE)
rownames(rowdata) <- rowdata$ensembl_gene_id
# Sort to match the order in the original data
rowdata <- rowdata[rownames(droplet_sce),]
## Check if  order is correct after merge;
stopifnot(all(rownames(rowdata) == rownames(droplet_sce)))
## add to the SingleCellExperiment object
rowData(droplet_sce) <- rowdata
```

For the remaining analysis, we will only focus on the 
`r format(sum(rowData(droplet_sce)$gene_biotype == "protein_coding", na.rm=TRUE), big.mark=",")`
protein coding genes that are contained in the data. These are selected below. 

```{r}
ind_pc <- which(rowData(droplet_sce)$gene_biotype == "protein_coding")
droplet_sce <- droplet_sce[ind_pc, ]
```

## QC and exploratory data analysis

In order to perform quality control (QC), we calculate some cell-specific
summary statistics: the total number of counts per cell (`sum`) and the total 
number of detected genes per cell (`detected`). These will be added in the
`colData` slot using the `addPerCellQC()` function from the `r Biocpkg("scater")`
package. 

```{r add-cellqc-metrics}
droplet_sce <- addPerCellQC(droplet_sce)
```

The distribution of these metrics per each cluster can be visualised as follows:

```{r cellqc-metrics, fig.cap="Cell-level QC metrics. For each cell cluster: distribution of the total number of UMIs per cell (left) and the total number of detected genes per cell (right)."}
p_cell_qc1 <- plotColData(droplet_sce, y = "sum", x = "subCellType") 
p_cell_qc2 <- plotColData(droplet_sce, y = "detected", x = "subCellType")
p_cell_qc1 + p_cell_qc2
```
As seen in Figure \ref(fig:cellqc-metrics), the  `presomiticMesoderm.b` cluster 
is characterised by a substantially larger number of UMIs and detected genes. 
These have been highlighted by the authors as potential doublets. As in the 
original publication, these will be removed from subsequent analyses. We also 
remove cells with more 25,000 UMIs. 

```{r drop-filt}
ind_retain <- colData(droplet_sce)$subCellType != "presomiticMesoderm.b" &
  colData(droplet_sce)$sum <= 25000
droplet_sce <- droplet_sce[, ind_retain]
```

After applying this filter, the dataset contains `r ncol(droplet_sce)` cells:
`r sum(colData(droplet_sce)$subCellType == "presomiticMesoderm.a")` in the 
presomitic Mesoderm group and 
`r sum(colData(droplet_sce)$subCellType == "somiticMesoderm")` in the somitic
Mesoderm group. 

<!--- We removed all cells that expressed <1,000 genes or that had >3% of their 
transcripts mapped to mitochondrial genes. We further removed any cells that 
expressed both Xist and any of Kdm5d, Eif2s3y, Gm29650, Uty or Ddx3y (genes in 
the Y chromosome), as these are probably doublets. ---> 

To further explore the underlying structure of the data, we perform global
scaling normalisation using `r Biocpkg("scran")` and
principal component analysis (PCA) of log-transformed normalised expression 
counts using `r Biocpkg("scater")`.
As seen in Figure \@ref(fig:pca-visualisation-stimulus-batch), this analysis 
suggests the absence of strong batch effects.
It should be noted that the estimation of global scaling normalisation factors 
using `r Biocpkg("scran")` is not strictly necessary in the 
`r Biocpkg("BASiCS")` workflow. 
Here, we only use it as part of the exploratory data analysis. 
Moreover, count-based models for dimensionality reduction 
(e.g. [@Townes2019;@Lopez2018]) could be used as an alternative to PCA, 
removing the need for log normalisation.

```{r pca-visualisation-mesoderm-batch, fig.cap="First two principal components of log-transformed expression counts after scran normalisation. Colour indicates the experimental condition (left), the number of detected genes (middle) and animal of origin (right) for each cell."}
## Global scaling normalisation 
droplet_sce <- computeSumFactors(droplet_sce, colData(droplet_sce)$subCellType)
## Obtain log-normalised expression counts
droplet_sce <- logNormCounts(droplet_sce)
## Run PCA
droplet_sce <- runPCA(droplet_sce)
## Plot first two PCs colour-coded according to cell-level metadata
p_type <- plotPCA(
    droplet_sce,
    ncomponents = c(2, 1),
    colour_by = "subCellType"
  ) +
  theme(legend.position = "bottom")
p_detected <- plotPCA(
    droplet_sce,
    ncomponents = c(2, 1),
    colour_by = "detected"
  ) +
  theme(legend.position = "bottom")
p_batch <- plotPCA(
    droplet_sce,
    ncomponents = c(2, 1),
    colour_by = "sample"
  ) +
  theme(legend.position = "bottom")
p_type + p_detected  + p_batch 
```
As seen inn Figure \ref(fig:pca-visualisation-mesoderm-batch), the first 
principal component (which explains approximately 9\%
of the total variability) captures the presomitic versus somitic Mesoderm
structure of the data. The second principal component (which explains 
approximately 3\% of the total variability) is correlated with the number of
detected features per cell. We do not observe substantial batch effects
driven by the animal of origin from which cells were collected. 

In addition to cell-specific QC, we also recommend a gene filtering 
step prior to using `r Biocpkg("BASiCS")`. 
The purpose of this filter is to remove lowly expressed genes that were largely
undetected through sequencing, making reliable variability estimates difficult
to obtain.
Here, we remove genes that are not detected in at least 20 cells across all 
cells. This is to ensure a reliable estimate of variability, roughly in line 
with the sample size requirements for the negative binomial distribution 
outlined in @Lloyd-Smith2007.


```{r gene-selection, fig.cap="Average UMI-count for each gene is plotted against the number of cells in which that gene was detected. Dashed grey lines are shown at the thresholds below which genes are removed."}
## Calculate gene-level QC metrics using scater
droplet_sce <- addPerFeatureQC(droplet_sce, exprs_values = "counts")
## Remove genes with zero total counts across all cells
droplet_sce <- droplet_sce[rowData(droplet_sce)$detected != 0, ]

## Transform "detected" into number of cells and define inclusion criteria
rowData(droplet_sce)$detected_cells <-
  rowData(droplet_sce)$detected * ncol(droplet_sce) / 100

## Set detection threshold to detecting expression in at least 20 cells
detected_threshold <- 20
rowData(droplet_sce)$include_gene <- 
  rowData(droplet_sce)$detected_cells >= detected_threshold

## Plot gene-level metrics, highlighting those included for our analysis
plotRowData(droplet_sce,
    x = "detected_cells",
    y = "mean",
    colour_by = "include_gene"
  ) +
  xlab("Number of cells in which expression was detected") +
  ylab("Average number of read counts across all cells") +
  scale_x_log10() +
  scale_y_log10() +
  theme(legend.position = "bottom") +
  geom_vline(
    xintercept = detected_threshold,
    linetype = "dashed",
    col = "grey60"
  )

## Apply gene filter
droplet_sce <- droplet_sce[rowData(droplet_sce)$include_gene, ]
```

The final dataset used in subsequent analyses contains 
`r ncol(droplet_sce)` cells and `r nrow(droplet_sce)` genes.
As the analysis will be carried out separately for pre-somitic and somitic
mesoderm cells, we create separate `SingleCellExperiment` object for each 
group of cells. These datasets contain 
`r sum(colData(droplet_sce)$cellType == "presomiticMesoderm")` and 
`r sum(colData(droplet_sce)$cellType == "somiticMesoderm")`, respectively. 

```{r separate-SCE}
sce_psm <- droplet_sce[, colData(droplet_sce)$cellType == "presomiticMesoderm"]
sce_sm <- droplet_sce[, colData(droplet_sce)$cellType == "somiticMesoderm"]
```

These are then stored for their use in the main manuscript.

```{r}
saveRDS(sce_psm, "rds/sce_psm.Rds")
saveRDS(sce_sm, "rds/sce_sm.Rds")
```

## BASiCS with and without spike-ins

TODO: write a basic intro here.

# Data pre-processing and quality control steps

## Downloading the data

The matrix with raw read counts can be obtained from ArrayExpress under the 
accession number 
[E-MTAB-4888](https://www.ebi.ac.uk/arrayexpress/experiments/E-MTAB-4888/).
In the matrix, column names contain library identifiers and row names
display Ensembl gene identifiers.

```{r naive-data}
if (!file.exists("downloads/")) {
  dir.create("downloads", showWarnings = FALSE)
}
if (!dir.exists("downloads/MarioniLab-ImmuneAging2017-5b665d1")) {
  website <- "https://zenodo.org/"
  folder <- "record/3522970/files/MarioniLab/"
  file <- "ImmuneAging2017-1.0.0.zip"
  download.file(
    paste0(website, folder, file),
    destfile = "downloads/ImmuneAging2017-1.0.0.zip"
  )
  unzip("downloads/ImmuneAging2017-1.0.0.zip", exdir = "downloads")
}
cd4_raw <- read.table(
  "downloads/MarioniLab-ImmuneAging2017-5b665d1/Data/raw_data.txt",
  header = TRUE,
  sep = "\t"
)
cd4_raw <- as.matrix(cd4_raw)
```

The input matrix contains data for `r format(ncol(cd4_raw), big.mark=",")`
cells and `r format(nrow(cd4_raw), big.mark=",")`
genes, including `r sum(grepl("ERCC", rownames(cd4_raw)))` ERCC spike-ins.

Information about experimental conditions and other metadata is available 
under the same accession number. 

```{r selecting-serum-cells}
cd4_metadata <- read_xlsx(
  "downloads/MarioniLab-ImmuneAging2017-5b665d1/Data/metadata.xlsx"
)
cd4_metadata <- as.data.frame(cd4_metadata)

## Save sample identifiers as rownames
rownames(cd4_metadata) <- cd4_metadata$CellName
```

The columns in the metadata file contain library identifiers (`X`), strain 
information (`Strain`; *Mus musculus castaneus* or *Mus musculus domesticus*), 
the age of the animals (`Age`; young or old), stimulation state of the cells 
(`Stimulus`; naive or activated), batch information (`Individuals`; associated 
to different mice), and cell type information (`Celltype`; via FACS or MACS 
purification). 

Here, we convert the data and metadata described above into a 
`SingleCellExperiment` object. 
For this purpose, we first separate the input matrix of expression counts into
two matrices associated to intrinsic genes and external spike-ins, respectively. 
Within the `SingleCellExperiment` object, the latter is stored separately
as an *alternative experiment*. For more details on the alternative experiment
slot, see `help("altExp")`.


```{r CD4-SCE-object}
## Separate intrinsic from ERCC counts
bio_counts <- cd4_raw[!grepl("ERCC", rownames(cd4_raw)), ]
spike_counts <- cd4_raw[grepl("ERCC", rownames(cd4_raw)), ]
## Generate the SingleCellExperiment object
sce_cd4_all <- SingleCellExperiment(
  assays = list(counts = bio_counts),
  colData = cd4_metadata[colnames(cd4_raw), ]
)
## Add read-counts for spike-ins as an alternative experiment
altExp(sce_cd4_all, "spike-ins") <- SummarizedExperiment(
  assays = list(counts = spike_counts)
)
```

Hereafter, our analysis focuses on naive CD4^+^ T cells in the presence and
absence of stimulation using plate-bound antibodies,+ obtained 
from young *Mus musculus domesticus* animals, and purified using MACS-based cell
sorting. 
Thus, we subset the `SingleCellExperiment` object to these 
`r sum(sce_cd4_all$Strain == "Mus musculus domesticus" & sce_cd4_all$Age == "Young" & sce_cd4_all$Celltype == "MACS-purified Naive")` cells.

```{r naive-activated-CD4-SCE-object}
ind_select <- sce_cd4_all$Strain == "Mus musculus domesticus" &
  sce_cd4_all$Age == "Young" &
  sce_cd4_all$Celltype == "MACS-purified Naive"
sce_naive_active <- sce_cd4_all[, ind_select]
sce_naive_active
```


## Annotation

Input data was annotated using Ensembl gene identifiers. 
To facilitate interpretation, it is often useful to obtain a mapping from 
Ensembl gene IDs to gene symbols using the BioMart suite
([http://www.biomart.org](http://www.biomart.org)) via the 
Bioconductor package `r Biocpkg("biomaRt")` [@Durinck2009].
This can also be used to obtain gene-pathways mappings and other
metadata (e.g. gene length), useful for performing functional analysis
of gene sets identified in downstream analyses.

```{r read-gene-symbols}
genenames <- readRDS("rds/genenames.Rds")
```

We add this information as `rowData` within the `SingleCellExperiment` 
object created above. 

```{r}
## Merge biomaRt annotation
rowdata <- data.frame(ensembl_gene_id = rownames(sce_naive_active))
rowdata <- merge(rowdata, genenames, by = "ensembl_gene_id", all.x = TRUE)
rownames(rowdata) <- rowdata$ensembl_gene_id
## Check if  order is correct after merge;
stopifnot(all(rownames(rowdata) == rownames(sce_naive_active)))
## add to the SingleCellExperiment object
rowData(sce_naive_active) <- rowdata
```

For the remaining analysis, we will only focus on the 
`r format(sum(rowData(sce_naive_active)$gene_biotype == "protein_coding", na.rm=TRUE), big.mark=",")`
protein coding genes that are contained in the data. These are selected below. 

```{r}
protein_coding <- which(
  rowData(sce_naive_active)$gene_biotype == "protein_coding"
)
sce_naive_active <- sce_naive_active[protein_coding, ]
```

## QC and exploratory data analysis

The data available at 
[E-MTAB-4888](https://www.ebi.ac.uk/arrayexpress/experiments/E-MTAB-4888/) have
been already filtered to remove poor quality samples.
The QC applied in [@Martinez-jimenez2017] removed cells with: (i) fewer 
than 1,000,000 total reads, (ii) less than 20% of reads mapped to 
endogenous genes, (iii) less than 1,250 or more than 3,000 detected genes and 
(iv) more than 10% or fewer than 0.5% of reads mapped to mitochondrial genes.
We include visualisations of these measures here; we also include another 
widely used QC diagnostic plot that compares the total 
number (or fraction) of spike-in counts versus the total number (or fraction) of 
endogeneous counts.
In such a plot, low quality samples are characterised by a high fraction of 
spike-in counts and a low fraction of endogeneous counts 
(see Figure \@ref(fig:PerCellQC)).

```{r PerCellQC, fig.cap="Cell-level QC metrics. The total number of endogenous read-counts (excludes non-mapped and intronic reads) is plotted against the total number of detected genes (left) and the total number of spike-in read-counts (right)."}
sce_naive_active <- addPerCellQC(sce_naive_active, use_altexps = TRUE)
p_cell_qc1 <- plotColData(
  sce_naive_active,
  x = "sum",
  y = "detected") +
  xlab("Total endogenous reads per cell") +
  ylab("Number of detected genes per cell") +
  theme(axis.text.x = element_text(hjust = 1, angle = 45))
p_cell_qc2 <- plotColData(
  sce_naive_active,
  x = "sum",
  y = "altexps_spike-ins_sum") +
  xlab("Total endogenous reads per cell") +
  ylab("Total spike-in reads per cell") +
  theme(axis.text.x = element_text(hjust = 1, angle = 45))
p_cell_qc1 + p_cell_qc2
```

We can also visualise these metrics with respect to cell-level metadata, such
as the experimental conditions (active vs unstimulated) and the different mice 
from which cells were collected 
(see Figure \@ref(fig:experimental-condition-batch)). 

```{r experimental-condition-batch, fig.cap="Cell-level QC metrics according to cell-level metadata. The total number of endogenous reads (excludes non-mapped and intronic reads) is plotted against the total number of detected genes. Colour indicates the experimental condition (left) and animal of origin (right) for each cell."}
p_stimulus <- plotColData(
    sce_naive_active,
    x = "sum",
    y = "detected",
    colour_by = "Stimulus"
  ) +
  xlab("Total endogenous reads per cell") +
  ylab("Number of detected genes per cell") +
  theme(
    legend.position = "bottom",
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
p_batch <- plotColData(
    sce_naive_active,
    x = "sum",
    y = "detected",
    colour_by = "Individuals"
  ) +
  xlab("Total endogenous reads per cell") +
  ylab("Number of detected genes per cell") +
  theme(
    legend.position = "bottom",
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
p_stimulus + p_batch
```

To further explore the underlying structure of the data, we perform global
scaling normalisation using `r Biocpkg("scran")` and
principal component analysis (PCA) of log-transformed normalised expression 
counts using `r Biocpkg("scater")`.
As seen in Figure \@ref(fig:pca-visualisation-stimulus-batch), this analysis 
suggests the absence of strong batch effects.
It should be noted that the estimation of global scaling normalisation factors 
using `r Biocpkg("scran")` is not strictly necessary in the 
`r Biocpkg("BASiCS")` workflow. 
Here, we only use it as part of the exploratory data analysis. 
Moreover, count-based models for dimensionality reduction 
(e.g. [@Townes2019;@Lopez2018]) could be used as an alternative to PCA, 
removing the need for log normalisation.

```{r pca-visualisation-stimulus-batch, fig.cap="First two principal components of log-transformed expression counts after scran normalisation. Colour indicates the experimental condition (left) and animal of origin (right) for each cell."}
## Global scaling normalisation + log tranformation + PCA
sce_naive_active <- computeSumFactors(sce_naive_active)
sce_naive_active <- logNormCounts(sce_naive_active)
sce_naive_active <- runPCA(sce_naive_active)
p_stimulus <- plotPCA(sce_naive_active, colour_by = "Stimulus") +
  theme(legend.position = "bottom")
p_batch <- plotPCA(sce_naive_active, colour_by = "Individuals") +
  theme(legend.position = "bottom")
p_stimulus + p_batch
```

In addition to cell-specific QC, we also recommend a gene filtering 
step prior to using `r Biocpkg("BASiCS")`. 
The purpose of this filter is to remove lowly expressed genes that were largely
undetected through sequencing, making reliable variability estimates difficult
to obtain.
Here, we remove genes that are not detected in at least 20 cells across both
conditions, or that have an average read count below 1. 
These thresholds can vary across datasets and should be 
informed by gene-specific QC metrics such as those shown in
Figure \@ref(fig:gene-selection) as well as prior knowledge about the cell types
and conditions being studied, where available.

```{r cd4-gene-selection, fig.cap="Average read-count for each gene is plotted against the number of cells in which that gene was detected. Dashed grey lines are shown at the thresholds below which genes are removed."}
sce_naive_active <- addPerFeatureQC(sce_naive_active, exprs_values = "counts")
## Remove genes with zero total counts across all cells
sce_naive_active <- sce_naive_active[rowData(sce_naive_active)$detected != 0, ]
## Transform "detected" into number of cells and define inclusion criteria
rowData(sce_naive_active)$detected_cells <-
  rowData(sce_naive_active)$detected * ncol(sce_naive_active) / 100
detected_threshold <- 20
mean_threshold <- 1
include_gene <- rowData(sce_naive_active)$mean >= mean_threshold &
  rowData(sce_naive_active)$detected_cells >= detected_threshold
rowData(sce_naive_active)$include_gene <- include_gene
plotRowData(
    sce_naive_active,
    x = "detected_cells",
    y = "mean",
    colour_by = "include_gene"
  ) +
  xlab("Number of cells in which expression was detected") +
  ylab("Average number of read counts across all cells") +
  scale_x_log10() +
  scale_y_log10() +
  theme(legend.position = "bottom") +
  geom_vline(
    xintercept = detected_threshold,
    linetype = "dashed",
    col = "grey60"
  ) +
  geom_hline(
    yintercept = mean_threshold,
    linetype = "dashed",
    col = "grey60"
  )
## Apply gene filter
sce_naive_active <- sce_naive_active[rowData(sce_naive_active)$include_gene, ]
```

Subsequently, we also require users to remove spike-in molecules that were not 
captured through sequencing. We do this separately for naive and active
cells.

```{r spike-ins-present}
ind_active <- sce_naive_active$Stimulus == "Active"
ind_naive <- sce_naive_active$Stimulus == "Unstimulated"
spikes <- assay(altExp(sce_naive_active))
detected_spikes_active <- rowSums(spikes[, ind_active] > 0) > 0
detected_spikes_naive <- rowSums(spikes[, ind_naive] > 0) > 0
detected_spikes <- detected_spikes_naive & detected_spikes_active
altExp(sce_naive_active) <- altExp(sce_naive_active)[detected_spikes, ]
```

## Spike-in concentrations

If spike-ins will be used to aid data normalisation and technical noise 
quantification, `r Biocpkg("BASiCS")` also requires the number of spike-in 
molecules that were added to each well. 
For each spike-in $i$, this corresponds to:

$$ \mu_{i} = C_i \times 10^{-18} \times (6.022 \times 10^{23}) 
\times V \times D \hspace{0.5cm} \mbox{where,} $$

- $C_i$ is the concentration for the spike-in $i$ (measured in $aM\mu{}l^{-1}$),
- $V$ is the volume added into each well (measure in $nl$) and
- $D$ is a dilution factor. 

The remaining factors in the equation above are unit conversion constants
(e.g. from moles to molecules). 
For the CD4^+^ T cell data, the authors added a 1:50,000 dilution of the ERCC 
spike-in mix 1 and a volume of $9nl$ was added into each well (see [https://www.fluidigm.com/faq/ifc-9](https://www.fluidigm.com/faq/ifc-9)).
Finally, input concentrations $C_i$ can be downloaded with the Bioconductor
package `r Biocpkg("scRNAseq")`, using the function
`ERCCSpikeInConcentrations()`.

Based on this information, we can use `BASiCS_CalculateERCC` to calculate
the amount of spike-in molecules in each well:

```{r, ercc-mul}
ERCC_count <- BASiCS_CalculateERCC(Mix = 1, DilutionFactor = 1/50000, VolumePerCell = 0.009)
```

To update the `sce_naive` and `sce_active` objects, the user must create a 
`data.frame` whose first column contains the spike-in labels (e.g. ERCC-00130) 
and whose second column contains the number of molecules calculated above.
We add this as row metadata for `altExp(sce_naive)` and 
`altExp(sce_active)`. 

```{r spike-info}
SpikeInput <- data.frame(
  Names = names(ERCC_count),
  count = ERCC_count
)
## Exclude spike-ins not included in the input SingleCellExperiment objects
ind_match <- match(rownames(altExp(sce_naive_active)), SpikeInput$Names)
SpikeInput <- SpikeInput[ind_match , ]
## Add as metadata
rowData(altExp(sce_naive_active)) <- SpikeInput
```

The final dataset used in subsequent analyses contains 
`r ncol(sce_naive_active)` cells, `r nrow(sce_naive_active)` genes and 
`r nrow(altExp(sce_naive_active))` spike-ins.

```{r save-SCE}
saveRDS(sce_naive_active, "rds/sce_naive_active.Rds")
ind_active <- sce_naive_active$Stimulus == "Active"
ind_naive <- sce_naive_active$Stimulus == "Unstimulated"
sce_naive <- sce_naive_active[, ind_naive]
sce_active <- sce_naive_active[, ind_active]
```



### Inference using BASiCS with spike-ins

```{r, eval = FALSE}
set.seed(42)
chain_naive <- BASiCS_MCMC(
  Data = sce_naive,
  PrintProgress = TRUE,
  N = 40000,
  Thin = 20,
  Burn = 20000,
  Regression = TRUE,
  Threads = 4,
  StoreChains = TRUE,
  StoreDir = "rds/",
  RunName = "naive",
  WithSpikes = FALSE
)
set.seed(43)
chain_active <- BASiCS_MCMC(
  Data = sce_active,
  PrintProgress = TRUE,
  N = 40000,
  Thin = 20,
  Burn = 20000,
  Regression = TRUE,
  Threads = 4,
  StoreChains = TRUE,
  StoreDir = "rds/",
  RunName = "active",
  WithSpikes = FALSE
)
```

### Inference using BASiCS without spike-ins


`r Biocpkg("BASiCS")`, when using spike-in molecules, uses
spike-ins as a reference in order to aid normalisation,
based on the assumption that the original quantity of spike-in molecules was
approximately equal in each well. @Eling2018 introduced a novel method of
inferring gene expression profiles using `r Biocpkg("BASiCS")` without
relying on spike-ins to quantify technical noise. This is useful
for droplet-based scRNAseq protocols, given that it is not possible to ensure
that each droplet contains a specified quantity of spike-in molecules.
In this horizontal integration framework, technical variation is quantified
using replication [@Carroll2005]. In the absence of true technical replicates,
we assume that population-level characteristics of the cells are replicated
using appropriate experimental design. This requires that cells from the same
population have been randomly allocated to different batches. Given appropriate
experimental design, `r Biocpkg("BASiCS")` assumes that biological effects
are shared across batches, while technical variation leads to spurious
differences between cells in different batches.

Using `r Biocpkg("BASiCS")` without spike-ins is very similar to using it
with spike-ins. We will demonstrate using the
naive cells. However, first, we must ensure that a `BatchInfo` field is present
in the `SingleCellExperiment` used as input. In this case we use individual
of origin as the batch vector.

```{r MCMC-naive-nospikes, eval = FALSE}
set.seed(44)
chain_naive_nospikes <- BASiCS_MCMC(
  Data = sce_naive,
  PrintProgress = TRUE,
  N = 40000,
  Thin = 20,
  Burn = 20000,
  Regression = TRUE,
  Threads = 4,
  StoreChains = TRUE,
  StoreDir = "rds/",
  RunName = "naive_nospikes",
  WithSpikes = FALSE
)
```

This MCMC chain and the chains run using spike-in data each took approximately
5 hours using an Intel 4770k processor with four threads.
As before, for convenience we provide a completed version of this chain at
[https://doi.org/10.5281/zenodo.5243265](https://doi.org/10.5281/zenodo.5243265).

```{r download-chain-naive-nospikes}
chains_website <- "https://git.ecdf.ed.ac.uk/vallejosgroup/basicsworkflow2020/-/raw/master/"
# chains_website <- "https://zenodo.org/record/5243265/files/"
if (!file.exists("rds/chain_naive_nospikes.Rds")) {
  download.file(
    paste0(chains_website, "rds/chain_naive_nospikes.Rds"),
    destfile = "rds/chain_naive_nospikes.Rds"
  )
}
chain_naive_nospikes <- readRDS("rds/chain_naive_nospikes.Rds")
```

The resulting `BASiCS_Chain` object produced using this horizontal integration
framework is functionally similar to one produced using the vertical integration
framework. It can be used in place of the `BASiCS_Chain` objects produced using
the vertical integration approach, as described above.

```{r diag-plot-naive-nospikes, fig.cap="Effective sample size is plotted against parameter estimates for mean and over-dispersion parameters using the horizontal integration approach. Colour indicates local density, with lighter colours representing higher density.", eval = FALSE, echo = FALSE}
As discussed previously, it is important to ensure that an adequate effective
sample size has been generated using MCMC.
Figure \@ref(fig:diag-plot-naive-nospikes) shows that the sampler has generated
a large effective sample size for most parameters, similar to the case when we 
 included spike-in molecules.
mu_ess2 <- BASiCS_DiagPlot(chain_naive_nospikes, Param = "mu") +
  theme(legend.position = "bottom")
delta_ess2 <- BASiCS_DiagPlot(chain_naive_nospikes, Param = "delta") +
  theme(legend.position = "bottom")
multiplot(mu_ess2, delta_ess2, cols = 2)
```

### Comparison of parameter estimates with and without spike-ins

Under the horizontal integration approach described above,
the scale of mean expression parameters and global scaling factors is not 
jointly identifiable, in that a global shift in mean expression parameters 
could be exactly offset by an equivalent shift in cell-specific normalisation
parameters. Therefore, the geometric mean of the mean expression parameters
is fixed to a constant value. Relative expression level estimates are broadly 
consistent between the horizontal and vertical integration approaches; 
however there may be a global difference in mean expression estimates, as shown
in Figure \@ref(fig:offset). It is important to remove this global scale offset
before performing comparative analyses. This is performed by default in
`BASiCS_TestDE`, but can be performed manually using `BASiCS_CorrectOffset`.

```{r download-chains}
options(timeout = 1000)
if (!file.exists("rds/chain_naive.Rds")) {
  download.file(
    paste0(chains_website, "chain_naive.Rds"),
    destfile = "rds/chain_naive.Rds"
  )
}
if (!file.exists("rds/chain_active.Rds")) {
  download.file(
    paste0(chains_website, "chain_active.Rds"),
    destfile = "rds/chain_active.Rds"
  )
}
chain_naive <- readRDS("rds/chain_naive.Rds")
chain_active <- readRDS("rds/chain_active.Rds")
```

```{r offset, fig.cap="Distribution of mean expression values before and after correcting the global difference in scale."}
BASiCS_PlotOffset(chain_naive_nospikes, chain_naive,
  GroupLabel1 = "No spike-ins", GroupLabel2 = "Spike-ins",
  Type = "before-after")
offset <- BASiCS_CorrectOffset(chain_naive_nospikes, chain_naive)
chain_naive_nospikes_offset <- offset$Chain
chain_naive_nospikes_offset
```

A number of genes have very low expression estimates in the naive population,
due to the fact that they each have zero read counts across the entire naive
population; we therefore remove these genes before making a comparison.
Following removal of the global offset, the mean expression 
and over-dispersion estimates obtained from each method are directly comparable.
As seen in Figures \@ref(fig:spike-comp)A and 
\@ref(fig:spike-comp)B, parameter point estimates from the two 
methods are highly correlated. There is a tail of non-expressed genes with
very low mean expression level as inferred without spike-ins, comprising
those genes with no measured expression across the entire population.

```{r spike-comp, fig.cap="Comparison of point estimates using spike-ins, and the same parameters estimated without using spike-ins for mean expression (A) and over-dispersion (B). A dashed red line indicates the identity line, $y=x$. Genes with zero counts across all cells were excluded from the plot of mean expression parameters."}
mu_spikes <- displayChainBASiCS(chain_naive)
mu_nospikes <- displayChainBASiCS(chain_naive_nospikes_offset)

# Remove genes with zero counts across all cells and calculate medians
ind_nonzero <- rowSums(counts(sce_naive)) != 0
mu_spikes <- colMedians(mu_spikes[, ind_nonzero])
mu_nospikes <- colMedians(mu_nospikes[, ind_nonzero])

g1 <- ggplot() +
  aes(mu_spikes, mu_nospikes) +
  geom_pointdensity(alpha = 0.7) +
  scale_colour_viridis(name = "Density") +
  scale_x_log10() +
  scale_y_log10() +
  geom_abline(
    colour = "firebrick",
    linetype = "dashed",
    slope = 1,
    intercept = 0
  ) +
  labs(
    x = "Mean expression\n(with spike-ins)",
    y = "Mean expression\n(without spike-ins)"
  ) +
  theme(
    legend.position = "bottom",
    legend.text = element_text(angle = 45, size = 8, hjust = 0.5, vjust = 0.5)
  )

delta_spikes <- displayChainBASiCS(chain_naive, Param = "delta")
delta_nospikes <- displayChainBASiCS(chain_naive_nospikes_offset, Param = "delta")

g2 <- ggplot() +
  aes(colMedians(delta_spikes), colMedians(delta_nospikes)) +
  geom_pointdensity(alpha = 0.7) +
  scale_colour_viridis(name = "Density") +
  scale_x_log10() +
  scale_y_log10() +
  geom_abline(
    colour = "firebrick",
    linetype = "dashed",
    slope = 1,
    intercept = 0
  ) +
  labs(
    x = "Over-dispersion\n(with spike-ins)",
    y = "Over-dispersion\n(without spike-ins)"
  ) +
  theme(
    legend.position = "bottom",
    legend.text = element_text(angle = 45, size = 8, hjust = 0.5, vjust = 0.5)
  )
g1 + g2 + plot_annotation(tag_levels = "A")
```


# Trace plots

First, we will load `r Biocpkg("BASiCS")` and read in each MCMC chain.
While it is important to ensure the MCMC algorithm has converged and has
produced an adequate number of independent samples from the posterior
distribution, we will here assume this proceeds as in the MCMC diagnostics
section of the main manuscript.

```{r load-chain-naive}
chain_naive <- readRDS("rds/chain_naive.Rds")
chain_active <- readRDS("rds/chain_active.Rds")
chain_naive_nospikes <- readRDS("rds/chain_naive_nospikes.Rds")
```

## Global scaling normalisation using BASiCS

Global scaling normalisation techniques for scRNA-seq data can be used to infer
cell-specific scaling factors that capture systematic differences in the scale
of the expression counts across cells. 
Generally, these scaling factors capture both biological (e.g. mRNA content) and technical (e.g. sequencing depth) factors, but spike-in genes can be used to
disentangle these effects [@Vallejos2017]. 
For example, when spike-in genes are used to assist data normalisation, 
`r Biocpkg("BASiCS")` infers two separate sets of cell-specific parameters 
associated to these sources of variability: $\phi_j$ and $\nu_j$, respectively.  
Posterior estimates for these parameters ($\hat{\phi}_j$, $\hat{\nu}_j$) can be 
used to normalise the input data via the `BASiCS_DenoisedCounts` function. 
This function returns normalised expression values defined as
$$
  x^*_{ij} = \frac{ x_{ij} } {\hat{\phi}_j \hat{\nu}_j},
$$
where $x_{ij}$ denotes the observed expression count for gene $i$ in cell $j$.

```{r normalisation-naive}
## First remove global differences in scale for normalisation factors
offset_corrected <- BASiCS_CorrectOffset(chain_active, chain_naive)
norm_basics_naive <- BASiCS_DenoisedCounts(
  Data = sce_naive,
  Chain = chain_naive
)
norm_basics_active <- BASiCS_DenoisedCounts(
  Data = sce_active,
  Chain = offset_corrected$Chain
)
```

As shown in Figure \@ref(fig:normalisation-vs-scran), the capture efficiency and 
scaling factors ($\nu_j$ and $\phi_j$ respectively) inferred by 
`r Biocpkg("BASiCS")` are highly correlated with the global scaling 
normalisation factors inferred using the method of @Lun2016.
We will load `r Biocpkg("scran")` to estimate these size factors.
An advantage of jointly inferring normalisation and gene-specific expression
parameters, rather than using pre-computed
normalisation factors, is that in the former case, statistical uncertainty is 
propagated from normalisation to all downstream analyses.

```{r normalisation-vs-scran, fig.width=4, fig.height=4, fig.cap="Comparison of BASiCS inferred global scaling factors and those estimated by scran using the method of Lun et al. (2016)."}
## BASiCS scaling factors within each group of samples
basics_sf_naive <- colMedians(
  displayChainBASiCS(chain_naive, "phi")
  * displayChainBASiCS(chain_naive, "nu")
)
basics_sf_active <- colMedians(
  displayChainBASiCS(chain_active, "phi")
  * displayChainBASiCS(chain_active, "nu")
)
## scran scaling factors within each group of samples
scran_sf_naive <- calculateSumFactors(sce_naive)
scran_sf_active <- calculateSumFactors(sce_active)

ggplot() +
  aes(basics_sf_naive, scran_sf_naive) +
  geom_point() +
  labs(x = "BASiCS scaling factors", y = "scran scaling factors") +
  annotate(
    x = 14,
    y = 0.4,
    label = paste(
      "Pearson's r = ",
      round(
        cor(basics_sf_naive, scran_sf_naive),
        digits = 2
      )
    ),
    geom = "text",
    size = 4
  )
```
