---
title: "BASiCS workflow: a step-by-step analysis of expression variability using single cell RNA sequencing data"
author: 
  - name: Nils Eling
    affiliation: 
    - &EBI European Molecular Biology Laboratory, European Bioinformatics 
      Institute, Wellcome Trust Genome Campus, Hinxton, Cambridge CB10 1SD, UK
    - &CRUK Cancer Research UK Cambridge Institute, University of Cambridge, 
      Li Ka Shing Centre, Cambridge, CB2 0RE, UK
    - &UZH Department of Quantitative Biomedicine, University of Zurich,
      Winterthurerstrasse 190, CH-8057, Zurich, Switzerland
  - name: Alan O'Callaghan
    affiliation: 
    - &MRC MRC Human Genetics Unit, Institute of Genetics \& Molecular Medicine, 
      University of Edinburgh, Western General Hospital, Crewe Road, Edinburgh, 
      EH4 2XU, UK
    email: "a.b.o'callaghan@sms.ed.ac.uk"
  - name: John C. Marioni
    affiliation: 
    - *EBI
    - *CRUK
  - name: Catalina A. Vallejos
    affiliation: 
    - *MRC 
    - The Alan Turing Institute, British Library, 96 Euston Road, London, 
      NW1 2DB, UK
    email: catalina.vallejos@igmm.ed.ac.uk
abstract: |
  Cell-to-cell gene expression variability is an inherent feature of complex 
  biological systems, such as immunity and development. Single-cell RNA 
  sequencing is a powerful tool to quantify this heterogeneity, but it is prone 
  to strong technical noise. In this article, we describe a step-by-step 
  computational workflow which uses the BASiCS Bioconductor package to robustly 
  quantify expression variability within and between known groups of cells (such 
  as experimental conditions or cell types). BASiCS uses an integrated framework 
  for data normalisation, technical noise quantification and downstream 
  analyses, whilst propagating statistical uncertainty across these steps. 
  Within a single seemingly homogeneous cell population, BASiCS can identify 
  highly variable genes that exhibit strong heterogeneity as well as lowly 
  variable genes with stable expression. BASiCS also uses a probabilistic 
  decision rule to identify changes in expression variability between cell 
  populations, whilst avoiding confounding effects related to differences in 
  technical noise or in overall abundance. Using two publicly available 
  datasets, we guide users through a complete pipeline which includes 
  preliminary steps for quality control as well as data exploration 
  using the scater and scran Bioconductor packages. Data for the first case 
  study was generated using the Fluidigm\@ C1 system, in which extrinsic 
  spike-in RNA molecules were added as a control. The second dataset was 
  generated using a droplet-based system, for which spike-in RNA is not 
  available. This analysis provides an example, in which differential 
  variability testing reveals insights regarding a possible early cell fate 
  commitment process. The workflow is accompanied by a Docker image that 
  ensures the reproducibility of our results. 
keywords: Single-cell RNA sequencing, expression variability, 
  transcriptional noise, differential expression testing
bibliography: Workflow.bib
urlcolor: Orange
output:
  BiocWorkflowTools::f1000_article:
    fig_width: 6
    fig_height: 3
  BiocStyle::html_document:
    fig_width: 8
    fig_height: 4
---



```{r setup_knitr, include = FALSE, cache = FALSE}
library("BiocStyle")
## Decide whether to display parts for BioC (TRUE) or F1000 (FALSE)
on.bioc <- FALSE
library("knitr")
library("ggplot2")
theme_set(theme_bw())
# Use fig.width = 7 for html and fig.width = 6 for pdf
# fig.width <- ifelse(on.bioc, 10, 6)
if (knitr::is_html_output()) {
  out_width <- "700px"
  out_height <- "600px"
} else if (knitr::is_latex_output()) {
  out_width <- "2.5in"
  out_height <- "3.5in"
}
knitr::opts_chunk$set(
  warning = FALSE, message = FALSE, error = FALSE,
  cache = 2, cache.path = "cache_main/",
  # fig.pos = "h",
  fig.path = "figure/"
)
```


# Introduction

<!--- scRNA-seq and the different types of heterogeneity ---> 
<!--- Nils to revisit and add additional references if required --->
Single-cell RNA-sequencing (scRNA-seq) enables the study of genome-wide 
transcriptional heterogeneity in cell populations that is not 
captured by bulk experiments [@Stegle2015; @Prakadan2017; @Patange2018]. 
On the broadest level, this heterogeneity can reflect the presence of distinct 
cell subtypes or states. 
Alternatively, it can be due to gradual changes along biological processes, 
such as development and differentiation. 
Several clustering and pseudotime inference methods have been developed to
characterise these types of heterogeneity [@Kiselev2019; @Saelens2019].
However, there is a limited availability of computational tools tailored 
to study more subtle variability within seemingly homogeneous cell populations. 
This variability can reflect deterministic or stochastic events that regulate
gene expression and, among others, has been reported to increase prior to cell
fate decisions [@Mojtahedi2016] as well as during ageing [@Martinez-jimenez2017]. 

<!--- Describes aims for the workflow and introduces BASiCS ---> 
This article complements existing scRNA-seq workflows based on the Bioconductor 
ecosystem (e.g. [@Lun2016; @Kim2019]), providing a detailed framework for 
transcriptional variability analyses. 
Firstly, we briefly discuss the sources of variability that arise in scRNA-seq 
data and the strategies that have been designed to control or attenuate 
technical noise in these assays. 
Subsequently, we describe a step-by-step workflow which uses
`r Biocpkg("scater")` [@McCarthy2017] and `r Biocpkg("scran")` [@Lun2016] 
to perform quality control (QC) as well as initial exploratory analyses. 
To robustly quantify transcriptional variability we use `r Biocpkg("BASiCS")` 
[@Vallejos2015; @Vallejos2016; @Eling2017] --- a Bayesian hierarchical framework 
that jointly performs data normalisation, technical noise quantification and 
downstream analyses, whilst propagating statistical uncertainty across these 
steps. 
Our analysis pipeline includes practical guidance to assess the convergence of 
the Markov Chain Monte Carlo (MCMC) algorithm that is used to infer model 
parameters as well as recommendations to interpret and post-process the model 
outputs. 
Finally, through a case study in the context of immune cells, we illustrate
how `r Biocpkg("BASiCS")` can be used to identify highly and lowly variable
genes within a cell population, as well as to compare expression profiles
between experimental conditions or cell types. 

All source code used to generate the results presented in this article is 
available [on Github](https://github.com/VallejosGroup/BASiCSWorkflow).
To ensure the 
reproducibility of this workflow, the analysis environment and all software 
dependencies are provided as a Docker image [@Boettiger2015]. The image 
can be obtained from
[Docker Hub](https://hub.docker.com/repository/docker/alanocallaghan/bocker).

# Sources of variability in scRNA-seq data

Stochastic variability within a seemingly homogeneous cell population --- often 
referred to as transcriptional *noise* --- can arise from intrinsic and 
extrinsic sources [@Elowitz2002; @Eling2019]. 
Extrinsic noise refers to stochastic fluctuations induced by 
different dynamic cellular states (e.g. cell cycle, metabolism, 
intra/inter-cellular signalling) [@Zopf2013; @Iwamoto2016; @Kiviet2014]. 
In contrast, intrinsic noise arises from stochastic effects on biochemical 
processes such as transcription and translation [@Elowitz2002].
Intrinsic noise can be modulated by genetic and epigenetic modifications (such 
as mutations, histone modifications, CpG island length and nucleosome 
positioning) [@Eberwine2015; @Faure2017; @Morgan2018] and usually occurs 
at the gene level [@Elowitz2002]. 
Cell-to-cell gene expression variability estimates derived from scRNA-seq data 
capture a combination of these effects, as well as deterministic regulatory 
mechanisms [@Eling2019]. 
Moreover, these variability estimates can also be inflated by the technical 
noise that is typically observed in scRNA-seq data [@Brennecke2013].

<!--- Experimental strategies to tackle technical noise --->
Different strategies have been incorporated into scRNA-seq protocols to control 
or attenuate technical noise. 
For example, external RNA spike-in molecules (such as the set introduced by the 
External RNA Controls Consortium, ERCC [@Rna2005]) can be added to each cell’s 
lysate in a (theoretically) known fixed quantity.
Spike-ins can assist quality control steps [@McCarthy2017], data normalisation
[@Vallejos2017] and can be used to infer technical noise [@Brennecke2013].
Another strategy is to tag individual cDNA molecules using unique molecular 
identifiers (UMIs) before PCR amplification [@Islam2014]. 
Reads that contain the same UMI can be collapsed into a single molecule count,
attenuating technical variability associated to cell-to-cell differences
in amplification and sequencing depth (these technical biases are not fully 
removed unless sequencing to saturation [@Vallejos2017]). 
However, despite the benefits associated to the use of spike-ins and UMIs, 
these are not available for all scRNA-seq protocols [@Haque2017].


# Methods {#methods}

This step-by-step scRNA-seq workflow is primarily based on the Bioconductor 
package ecosystem [@Amezquita2019]. 
A graphical overview is provided in Figure \@ref(fig:overview) 
and its main components are described below. 

```{r overview, out.width=out_width, out.height=out_height, fig.cap = 'Graphical overview for the scRNA-seq analysis workflow described in this manuscript. Starting from a matrix of expression counts, we use the scater and scran Bioconductor packages to perform QC and initial exploratory analyses. To robustly quantify transcriptional heterogeneity within seemingly homogeneous cell populations, we apply the BASiCS Bioconductor package and  illustrate how BASiCS can be used to analyse a single or multiple pre-specified groups of cells.', echo=FALSE}
knitr::include_graphics("figure/Overview.svg")
```

## Input data

```{r}
library("SingleCellExperiment")
```

We use `r Biocpkg("SingleCellExperiment")` to convert an input
matrix of raw read-counts (molecule counts for UMI-based protocols) into a 
`SingleCellExperiment` object which can also store its associated 
metadata, such as gene- and cell-specific information. 
Moreover, when available, the same object can also store read-counts for 
spike-in molecules (see `altExp()`).
A major advantage of using a `SingleCellExperiment` object as the input for 
scRNA-seq analyses is the interoperability across a large number of 
Bioconductor packages [@Amezquita2019]. 

## Quality control and exploratory analysis

```{r}
library("scater")
library("scran")
```

An critical step in scRNA-seq analyses is QC, removing low quality samples that 
may distort downstream analyses. 
Among others, QC diagnostics can help to identify samples that contain broken 
cells, that are empty or that contain multiple cells [@Ilicic2016]. 
Moreover, lowly expressed genes for which less reliable information is 
available are typically also removed. 
The [*OSCA*](https://osca.bioconductor.org/) online book provides an extensive
overview on important aspects of how to perform QC of scRNA-seq data, including
exploratory analyses [@Amezquita2019].

Here, we use the `r Biocpkg("scater")` package [@McCarthy2017] to calculate 
QC metrics for each cell (e.g. total read-count) and gene (e.g. percentage of 
zeroes across all cells), respectively. 
Moreover, we use the visualisation tools implemented in `r Biocpkg("scater")` to 
explore the input dataset and its associated QC diagnostic metrics. 
For further data exploration we use the `r Biocpkg("scran")` package [@Lun2016].
`r Biocpkg("scran")` can perform *global scaling* normalisation, calculating 
cell-specific scaling factors that capture global differences in read-counts 
across cells (e.g. due to sequencing depth and PCR amplification) 
[@Lun2016pooling]. 
Moreover, `r Biocpkg("scran")` enables exploratory analyses of transcriptional
variability. 
For example, it can be used to infer an overall trend between mean expression 
and the squared coefficent of variation (CV^2^) for each gene. 
To derive variability estimates that are not confounded by this overall trend, 
`r Biocpkg("scran")` also defines gene-specific DM (distance to the mean) 
estimates as the distance between CV$^2$ and a rolling median along the range 
of mean expression values [@Kolodziejczyk2015cell].
DM estimates enable exploratory analyses of cell-to-cell heterogeneity, but a
measure of uncertainty is not readily available. As such, gene-specific 
downstream inference (e.g. differential variability testing) is precluded.

## BASiCS - Bayesian Analysis of Single Cell Sequencing data

```{r}
library("BASiCS")
```

The `r Biocpkg("BASiCS")` package uses a Bayesian hierarchical framework 
that borrows information across all genes and cells to robustly quantify 
transcriptional variability [@Vallejos2015BASiCS].
Similar to the approach adopted in `r Biocpkg("scran")`, `r Biocpkg("BASiCS")` 
infers cell-specific global scaling normalisation parameters. 
However, instead of inferring these as a pre-processing step, 
`r Biocpkg("BASiCS")` uses an integrated approach in which data normalisation 
and downstream analyses are performed simultaneously, thereby propagating
statistical uncertainty. 
To quantify technical noise, the original implementation of 
`r Biocpkg("BASiCS")` uses information from extrinsic spike-in molecules as 
control features, but the model has been extended to address situations in which
spike-ins are not available [@Eling2018].

`r Biocpkg("BASiCS")` summarises expression patterns through 
gene-specific *mean* ($\mu_i$) and *over-dispersion* ($\delta_i$) parameters. 
Mean parameters $\mu_i$ quantify the overall expression for each gene $i$
across the population of cells under study.
In contrast, $\delta_i$ captures the excess of variability that is observed with 
respect to what would be expected in a homogeneous cell population, after 
taking into account technical noise. 
This is used as a proxy to quantify transcriptional variability.
To account for the strong relationship that is typically observed 
between gene-specific mean expression and over-dispersion estimates, 
Eling *et al.*  [@Eling2018] recently introduced a joint prior specification for 
these parameters. 
This joint prior formulation has been observed to improve posterior inference
when the data is less informative (e.g. small sample size, lowly expressed 
genes) and can be used to derive gene-specific *residual over-dispersion* 
parameters $\epsilon_i$ that are not confounded by mean expression. 
Similar to DM values implemented in `r Biocpkg("scran")`, these are defined as 
deviations with respect to an overall regression trend that captures the 
relationship between mean and over-dispersion values.

Within a population of cells, `r Biocpkg("BASiCS")` decomposes the total 
observed variability in expression measurements into technical and biological 
components [@Vallejos2015]. 
This enables the identification of *highly variable genes* (HVGs) that capture
the major sources of heterogeneity within the analysed cells [@Brennecke2013]. 
HVG detection is often used as feature selection, to identify the input 
set of genes for subsequent analyses. 
`r Biocpkg("BASiCS")` can also highlight *lowly variable genes* (LVGs) that 
exhibit stable expression across the population of cells.
These may relate to essential cellular functions and can assist the development
of new data normalisation or integration strategies [@Lin2019].

`r Biocpkg("BASiCS")` also provides a probabilistic decision rule to 
perform differential expression analyses between two (or more) pre-specified 
groups of cells [@Vallejos2016; @Eling2018].
Whilst several differential expression tools have been proposed for scRNA-seq 
data (e.g. [@Kharchenko2014; @Finak2015]), some evidence suggests that 
these do not generally outperform popular bulk RNA-seq tools [@Soneson2018]. 
Moreover, most of these methods are only designed to uncover changes in overall
expression, ignoring the more complex patterns that can arise at the single cell 
level [@Lahnemann2020]. 
Instead, `r Biocpkg("BASiCS")` embraces the high granularity of scRNA-seq data,
uncovering changes in cell-to-cell expression variability that are not 
confounded by differences in technical noise or in overall expression. 

# Case study: analysis of naive CD4^+^ T cells {#Tcells}

As a case study, we use scRNA-seq data generated for CD4^+^ T cells
using the C1 Single-Cell Auto Prep System (Fluidigm^®^). 
Martinez-Jimenez _et al._ profiled naive (hereafter also referred to as 
unstimulated) and activated (3 hours using _in vitro_ antibody stimulation) 
CD4^+^ T cells from young and old animals across two mouse strains to study 
changes in expression variability during ageing and upon immune activation 
[@Martinez-jimenez2017].
They extracted naive or effector memory CD4^+^ T cells from spleens of young or 
old animals, obtaining purified populations using either magnetic-activated cell 
sorting (MACS) or fluorescence activated cell sorting (FACS).
External ERCC spike-in RNA [@Rna2005] was added to aid the quantification of 
technical variability across all cells and all experiments were performed in 
replicates (also referred to as batches) to control for batch effects.

## Obtaining the data

The matrix with raw read counts can be obtained from ArrayExpress under the 
accession number 
[E-MTAB-4888](https://www.ebi.ac.uk/arrayexpress/experiments/E-MTAB-4888/).
In the matrix, column names contain library identifiers and row names
display gene Ensembl identifiers. 

```{r naive-data}
if (!file.exists("downloads/raw_data.txt")) {
  website <- "https://www.ebi.ac.uk/arrayexpress/files/E-MTAB-4888/"
  file <- "E-MTAB-4888.processed.1.zip"
  destfile <- "downloads/raw_data.txt.zip"
  download.file(
    paste0(website, file),
    destfile = destfile
  )
  unzip("downloads/raw_data.txt.zip", exdir = "downloads")
  file.remove("downloads/raw_data.txt.zip")
}

# Read in raw data
CD4_raw <- read.table("downloads/raw_data.txt", header = TRUE, sep = "\t")
CD4_raw <- as.matrix(CD4_raw)
```

The input matrix contains data for `r format(ncol(CD4_raw), big.mark=",")`
cells and `r format(nrow(CD4_raw), big.mark=",")`
genes (including `r sum(grepl("ERCC", rownames(CD4_raw)))` ERCC spike-ins). 
Information about experimental conditions and other metadata is available 
under the same accession number. 

```{r selecting-serum-cells}
if (!file.exists("downloads/metadata_file.txt")) {
  website <- "https://www.ebi.ac.uk/arrayexpress/files/E-MTAB-4888"
  file <- "E-MTAB-4888.additional.1.zip"
  download.file(
    paste0(website, file),
    destfile = "downloads/metadata.txt.zip"
  )
  unzip("downloads/metadata.txt.zip", exdir = "downloads")
  file.remove("downloads/metadata.txt.zip")
}

CD4_metadata <- read.table(
  "downloads/metadata_file.txt",
  header = TRUE,
  sep = "\t"
)

# Save sample identifiers as rownames
rownames(CD4_metadata) <- CD4_metadata$X
```

The columns in the metadata file contain library identifiers (`X`), strain 
information (`Strain`; *Mus musculus castaneus* or *Mus musculus domesticus*), 
the age of the animals (`Age`; young or old), stimulation state of the cells 
(`Stimulus`; naive or activated), batch information (`Individuals`; associated 
to different mice), and cell type information (`Celltype`; via FACS or MACS 
purification). 

The data and metadata described above are then converted into a 
`SingleCellExperiment` object. 
For this purpose, we first separate the input matrix of expression counts into
separate matrices associated to intrinsic genes and external spike-ins. 
Within the `SingleCellExperiment` object, the latter is stored separately
through as an alternative experiment (see `?altExp`).

```{r CD4-SCE-object}
# Separate intrinsic from ERCC counts
bio_counts <- CD4_raw[!grepl("ERCC", rownames(CD4_raw)), ]
spike_counts <- CD4_raw[grepl("ERCC", rownames(CD4_raw)), ]
# Generate the SingleCellExperiment object
sce_CD4_all <- SingleCellExperiment(
  assays = list(counts = as.matrix(bio_counts)),
  colData = CD4_metadata[colnames(CD4_raw), ]
)
# Add read-counts for spike-ins as an alternative experiment
altExp(sce_CD4_all, "spike-ins") <- SummarizedExperiment(
  assays = list(counts = spike_counts)
)
```

Hereafter our analysis focuses on naive and activated CD4^+^ T cells obtained 
from young *Mus musculus domesticus* animals, purified using MACS-based cell sorting. 
Here, we extract these 
`r sum(sce_CD4_all$Strain == "Mus musculus domesticus" & sce_CD4_all$Age == "Young" & sce_CD4_all$Celltype == "MACS-purified Naive")` samples.

```{r naive-activated-CD4-SCE-object}
ind_select <- sce_CD4_all$Strain == "Mus musculus domesticus" &
  sce_CD4_all$Age == "Young" &
  sce_CD4_all$Celltype == "MACS-purified Naive"
sce_naive_active <- sce_CD4_all[, ind_select]
sce_naive_active
```

## QC and exploratory analysis

The data available at 
[E-MTAB-4888](https://www.ebi.ac.uk/arrayexpress/experiments/E-MTAB-4888/) have
been already filtered to remove poor quality samples.
The QC applied in [@Martinez-jimenez2017] removed cells with: (i) fewer 
than 1,000,000 total reads, (ii) less than 20% of reads mapped to 
endogenous genes, (iii) less than 1,250 or more than 3,000 detected genes and 
(iv) more than 10% or fewer than 0.5% of reads mapped to mitochondrial genes.
As an illustration, we visualise some of these metrics. 
We also include another widely used QC diagnostic plot which compares the total 
number (or fraction) of spike-in counts versus the total number (or fraction) of 
endogeneous counts.
In such a plot, low quality samples are characterised by a high fraction of 
spike-in counts and a low fraction of endogeneous counts 
(see Figure \@ref(fig:PerCellQC)).

```{r PerCellQC, fig.cap="Cell-level QC metrics. The total number of endogenous read-counts (excludes non-mapped and intronic reads) is plotted against the total number of detected genes (left) and the total number of spike-in read-counts (right)."}
# Calculate and plot per cell QC metrics
sce_naive_active <- addPerCellQC(sce_naive_active, use_altexps = TRUE)
p_cellQC1 <- plotColData(
  sce_naive_active, 
  x = "sum", 
  y = "detected") +
  xlab("Total engogenous reads per cell") +
  ylab("Number of detected genes per cell")
p_cellQC2 <- plotColData(
  sce_naive_active, 
  x = "sum", 
  y = "altexps_spike-ins_sum") +
  xlab("Total engogenous reads per cell") +
  ylab("Total spike-in reads per cell")
multiplot(p_cellQC1, p_cellQC2, cols = 2)
```

These metrics can also be visualised with respect to cell-level metadata, such
as the experimental conditions (active vs unstimulated) and the different mice 
from which cells were collected (see Figure \@ref(fig:experimental-condition-batch)). 

```{r experimental-condition-batch, fig.cap="Cell-level QC metrics according to cell-level metadata. The total number of endogenous reads (excludes non-mapped and intronic reads) is plotted against the total number of detected genes. Colour indicates the experimental condition (left) and animal of origin (right) for each cell."}
p_stimulus <- plotColData(
    sce_naive_active,
    x = "sum",
    y = "detected", 
    colour_by = "Stimulus") +
  xlab("Total engogenous reads per cell") +
  ylab("Number of detected genes per cell") +
  theme(
    legend.position = "bottom",
    axis.text.x = element_text(angle = 45, hjust = 1))
p_batch <- plotColData(
    sce_naive_active,
    x = "sum",
    y = "detected", 
    colour_by = "Individuals") +
  xlab("Total engogenous reads per cell") +
  ylab("Number of detected genes per cell") +
  theme(
    legend.position = "bottom",
    axis.text.x = element_text(angle = 45, hjust = 1))
multiplot(p_stimulus, p_batch, cols = 2)
```

To further explore the underlying structure of the data, we compute global
scaling normalisation factors using `r Biocpkg("scran")` and perform a 
principal component analysis (PCA) of log-transformed normalised expression 
counts using `r Biocpkg("scater")`. 
As seen in Figure \@ref(fig:pca-visualisation-stimulus-batch), this analysis 
suggests the absence of strong batch effects. 

```{r pca-visualisation-stimulus-batch, fig.cap="First two principal components of log-transformed expression counts after scran normalisation. Colour indicates the experimental condition (left) and animal of origin (right) for each cell."}
# Global scaling normalisation
sce_naive_active <- computeSumFactors(sce_naive_active)
sce_naive_active <- logNormCounts(sce_naive_active)
# PCA 
sce_naive_active <- runPCA(sce_naive_active)
p_stimulus <- plotPCA(sce_naive_active, colour_by = "Stimulus") +
  theme(legend.position = "bottom")
p_batch <- plotPCA(sce_naive_active, colour_by = "Individuals") +
  theme(legend.position = "bottom")
multiplot(p_stimulus, p_batch, cols = 2)
```

In addition to cell-specific QC, we also recommend to apply a gene filtering 
step prior to the use of `r Biocpkg("BASiCS")`. 
The purpose of this filter is to remove lowly expressed genes that were largely
undetected through sequencing and for which variability estimates are less 
reliable. 
Here, we remove all genes that are not detected in at least 5 cells across both 
experimental conditions or for which their average read count is below 1. 
These thresholds can vary across datasets and might be informed by gene-specific 
QC metrics such as those shown in \@ref(fig:gene-selection).

```{r gene-selection, fig.cap="Average read-count for each gene is plotted against the number of cells in which that gene was detected. Dashed grey lines are shown at the thresholds below which genes are removed."}
# Calculate per gene QC metrics
sce_naive_active <- addPerFeatureQC(sce_naive_active, exprs_values = "counts")
# Remove genes with zero total counts across all cells
sce_naive_active <- sce_naive_active[rowData(sce_naive_active)$detected != 0, ]
# Transform 'detected' metadata into number of cells 
rowData(sce_naive_active)$detected_cells <- 
  rowData(sce_naive_active)$detected * ncol(sce_naive_active) / 100
# Define inclusion criteria
rowData(sce_naive_active)$include_gene <- rowData(sce_naive_active)$mean >= 1 &
  rowData(sce_naive_active)$detected_cells >= 5
  
plotRowData(
    sce_naive_active,
    x = "detected_cells",
    y = "mean",
    colour_by = "include_gene") +
  xlab("Total engogenous reads per cell") +
  ylab("Number of detected genes per cell") +
  scale_x_log10() +
  scale_y_log10() +
  theme(
    legend.position = "bottom",
    axis.text.x = element_text(angle = 45, hjust = 1)) +
geom_vline(xintercept = 5, linetype = "dashed", col = "grey60") +
geom_hline(yintercept = 1, linetype = "dashed", col = "grey60")
# Apply inclusion criteria
sce_naive_active <- sce_naive_active[rowData(sce_naive_active)$include_gene, ]
```

Subsequently we also recommend to remove spike-in molecules that were not 
captured through sequencing. 

```{r spike-ins-present}
detected_spikes <- rowSums(assay(altExp(sce_naive_active)) > 0) > 0
altExp(sce_naive_active) <- altExp(sce_naive_active)[detected_spikes, ]
```

The final dataset used in subsequent analyses contains `r ncol(sce_naive_active)`
cells, `r nrow(sce_naive_active)` genes and `r nrow(altExp(sce_naive_active))` spike-ins. 

## BASiCS analysis - input data

Here, we apply the `r Biocpkg("BASiCS")` model separately to cells from each
experimental condition (`r sum(sce_naive_active$Stimulus == "Unstimulated")` 
naive and `r sum(sce_naive_active$Stimulus == "Active")` activated cells). 
Separate `SingleCellExperiment` objects are created for each group of cells. 

```{r SCE-separation}
sce_naive <- sce_naive_active[, sce_naive_active$Stimulus == "Unstimulated"]
sce_active <- sce_naive_active[, sce_naive_active$Stimulus == "Active"]
```

`r Biocpkg("BASiCS")` requires the user to update these objects with additional
information, using a specific format.
Firstly, if multiple batches of sequenced cells are available (e.g. multiple 
donors from which cells were extracted or multiple sequencing batches from the 
same experimental condition), this information must be included under the 
`BatchInfo` label as part of the cell-level metadata.

```{r BatchInfo}
colData(sce_naive)$BatchInfo <- colData(sce_naive)$Individuals
colData(sce_active)$BatchInfo <- colData(sce_active)$Individuals
```

If spike-ins will be used to aid data normalisation and technical noise 
quantification, `r Biocpkg("BASiCS")` also requires the number of spike-in 
molecules that were added to each well. 
For each spike-in gene $i$, this corresponds to:

$$ \mu_{i} = C_i \times 10^{-18} \times (6.022 \times 10^{23}) 
\times V \times D \hspace{0.5cm} \mbox{where,} $$

- $C_i$ is the concentration for the spike-in $i$ (measured in $aM\mu{}l^{-1}$),
- $V$ is the volume added into each well (measure in $nl$) and
- $D$ is a dilution factor. 

The remaining factors in the equation above are unit conversion constants (e.g.
from moles to molecules). 
For the CD4^+^ T cell data, the authors added a 1:50,000 dilution of the ERCC 
spike-in mix 1 and a volume of $9nl$ was added into each well (see [https://www.fluidigm.com/faq/ifc-9](https://www.fluidigm.com/faq/ifc-9)).
Finally, input concentrations $C_i$ can be downloaded from [https://assets.thermofisher.com/TFS-Assets/LSG/manuals](https://assets.thermofisher.com/TFS-Assets/LSG/manuals/cms_095046.txt)). 

```{r spike-in_download}
if (!file.exists("downloads/spike_info.txt")) {
  website <- "https://assets.thermofisher.com/TFS-Assets/LSG/manuals"
  file <- "cms_095046.txt"
  download.file(
    paste0(website, file),
    destfile = "downloads/spike_info.txt"
  )  
}

ERCC_conc <- read.table("downloads/spike_info.txt", sep = "\t", header = TRUE)
```

Based on this information, the calculation above proceeds as follows

```{r, ercc-mul}
# Moles per micro litre
ERCC_mmul <- ERCC_conc$concentration.in.Mix.1..attomoles.ul. * (10^(-18))
# Molecule count per micro litre (1 mole comprises 6.02214076 x 10^{23} molecules)
ERCC_countmul <- ERCC_mmul * (6.02214076 * (10^23))
# Application of the dilution factor (1:50,000)
ERCC_count <- ERCC_countmul / 50000
# Multiplying by the volume added into each well
ERCC_count_final <- ERCC_count * 0.009
```

To update the `sce_naive` and `sce_active` objects, the user must create a 
`data.frame` whose first column contains the labels associated to 
the spike-in molecule (e.g. ERCC-00130) and whose second column contains the 
input number of molecules calculated above.
This information is added as metadata for `altExp(sce_naive)` and 
`altExp(sce_active)`, respectively. 

```{r spike-info}
SpikeInput <- data.frame(
  Names = ERCC_conc$ERCC.ID,
  count = ERCC_count_final
)
# Exclude spike-ins not included in the input SingleCellExperiment objects
SpikeInput <- subset(SpikeInput, Names %in% rownames(altExp(sce_naive)))

metadata(sce_naive)$SpikeInput <- SpikeInput
metadata(sce_active)$SpikeInput <- SpikeInput
```

## BASiCS analysis - parameter estimation

Parameter estimation is implemented in the `BASiCS_MCMC` function using an 
adaptive Metropolis within Gibbs algorithm [@Roberts2009].
The primary inputs for `BASiCS_MCMC` correspond to:

- `Data`: a `SingleCellExperiment` object created as described in the previous
sections. 
- `N`: the total number of MCMC iterations
- `Thin`: thining period for output storage (only the `Thin`-th MCMC draw will be stored)
- `Burn`: the initial number of MCMC iterations to be discarded
- `Regression`: whether or not the join ()
- `WithSpikes`:

As a default, we recommend to use `Regression = TRUE` as we have observed that
the joint prior introduced by Eling *et al.* leads to m

Additional optional parameters can be used to store the generated output 
(`StoreChains`, `StoreDir`, `RunName`) and to monitor the progress of the 
algorithm (`PrintProgress`). 

<!--- whose convergence can
be assessed using the `BASiCS_DiagHist` and `BASiCS_DiagPlot` functions, 
among others. 
The output from `BASiCS_MCMC` is a `BASiCS_Chain` object, which can be used 
for further downstream analyses. In particular, `BASiCS_DetectHVG` and 
`BASiCS_DetectLVG` can be respectively used to identify highly and lowly 
variable genes within a cell population. Moreover, `BASiCS_TestDE` is used to
perform differential mean and variability analyses between groups of cells.--->



```{r}
library("ggplot2")
```



<!--- BASiCS downstream analyses between populations 
--->



