---
title: "BASiCS workflow: a step-by-step analysis of expression variability using single cell RNA sequencing data"
author: 
  - name: Alan O'Callaghan
    affiliation: 
    - &MRC MRC Human Genetics Unit, Institute of Genetics \& Cancer, 
      University of Edinburgh, Western General Hospital, Crewe Road, Edinburgh, 
      EH4 2XU, UK
    email: "alan.ocallaghan@outlook.com"
  - name: Nils Eling
    affiliation: 
    - &UZH Department of Quantitative Biomedicine, University of Zurich,
      Winterthurerstrasse 190, CH-8057, Zurich, Switzerland
    - &ETH Institute for Molecular Health Sciences, ETH Zurich,
      Otto-Stern Weg 7, 8093 Zurich, Switzerland
  - name: John C. Marioni
    affiliation: 
    - &EBI European Molecular Biology Laboratory, European Bioinformatics 
      Institute, Wellcome Trust Genome Campus, Hinxton, Cambridge CB10 1SD, UK
    - &CRUK Cancer Research UK Cambridge Institute, University of Cambridge, 
      Li Ka Shing Centre, Cambridge, CB2 0RE, UK
  - name: Catalina A. Vallejos
    affiliation: 
    - *MRC 
    - The Alan Turing Institute, British Library, 96 Euston Road, London, 
      NW1 2DB, UK
    email: catalina.vallejos@igmm.ed.ac.uk
abstract: |
  Cell-to-cell gene expression variability is an inherent feature of complex 
  biological systems, such as immunity and development. Single-cell RNA 
  sequencing is a powerful tool to quantify this heterogeneity, but it is prone 
  to technical noise, beyond what is observed in bulk-level assays.
  In this article, we describe a step-by-step 
  computational workflow that uses the BASiCS Bioconductor package to robustly 
  quantify expression variability within and between known groups of cells (such 
  as experimental conditions or cell types). BASiCS uses an integrated framework 
  for data normalisation, technical noise quantification and downstream 
  analyses, propagating statistical uncertainty across these steps. 
  Within a single seemingly homogeneous cell population, BASiCS can identify 
  highly variable genes that exhibit strong heterogeneity as well as lowly 
  variable genes with stable expression. BASiCS also uses a probabilistic 
  decision rule to identify changes in expression variability between cell 
  populations, whilst avoiding confounding effects related to differences in 
  technical noise or in overall abundance. Using a publicly available 
  dataset, we guide users through a complete pipeline that includes 
  preliminary steps for quality control, as well as data exploration 
  using the scater and scran Bioconductor packages. The workflow is accompanied 
  by a Docker image that ensures the reproducibility of our results. 
keywords: Single-cell RNA sequencing, expression variability, 
  transcriptional noise, differential expression testing
bibliography: Workflow.bib
urlcolor: Orange
output:
  BiocWorkflowTools::f1000_article:
    fig_width: 6
    fig_height: 3.5
---

```{r setup_knitr, include = FALSE, cache = FALSE}
library("BiocStyle")
## Decide whether to display parts for BioC (TRUE) or F1000 (FALSE)
on.bioc <- FALSE
library("knitr")
library("ggplot2")
theme_set(theme_bw())
## Use fig.width = 7 for html and fig.width = 6 for pdf
## fig.width <- ifelse(on.bioc, 10, 6)
if (knitr::is_html_output()) {
  out_width <- "700px"
  out_height <- "600px"
} else if (knitr::is_latex_output()) {
  out_width <- "2.5in"
  out_height <- "3.5in"
}
knitr::opts_chunk$set(
  warning = FALSE, message = FALSE, error = FALSE,
  cache = 2, cache.path = "cache_main/",
  # fig.pos = "h",
  fig.path = "figure/"
)
options(timeout=100)
```

# Introduction

<!--- scRNA-seq and the different types of heterogeneity ---> 
Single-cell RNA-sequencing (scRNA-seq) enables the study of genome-wide 
cell-to-cell transcriptional heterogeneity that is not captured by bulk 
experiments [@Stegle2015; @Prakadan2017; @Patange2018]. 
On the broadest level, this heterogeneity can reflect the presence of distinct 
cell subtypes or states. 
Alternatively, it can be due to gradual changes along biological processes, 
such as development and differentiation. 
Several clustering and pseudotime inference tools have been developed to
capture these types of heterogeneity [@Kiselev2019; @Saelens2019].
However, there is a limited availability of tools tailored 
to study more subtle variability within seemingly homogeneous cell populations. 
This variability can reflect deterministic or stochastic events that regulate 
gene expression and, among other settings, has been seen to increase prior to 
cell fate decisions [@Mojtahedi2016] and during ageing [@Martinez-jimenez2017]. 
Transcriptional variability has also been observed to differ from gene to gene 
and can be conserved across cell types and species [@Lin2019]. 

<!--- Narrow focus to transcriptional noise --->
Stochastic variability within a seemingly homogeneous cell population --- often 
referred to as transcriptional *noise* --- can arise from intrinsic and 
extrinsic sources [@Elowitz2002; @Eling2019]. 
Extrinsic noise refers to stochastic fluctuations induced by 
different dynamic cellular states (e.g. cell cycle, metabolism, 
intra/inter-cellular signalling) [@Zopf2013; @Iwamoto2016; @Kiviet2014]. 
In contrast, intrinsic noise arises from stochastic effects on biochemical 
processes such as transcription and translation [@Elowitz2002].
Intrinsic noise can be modulated by genetic and epigenetic modifications (such 
as mutations, histone modifications, CpG island length and nucleosome 
positioning) [@Eberwine2015; @Faure2017; @Morgan2018] and usually occurs 
at the gene level [@Elowitz2002]. 
Cell-to-cell gene expression variability estimates derived from scRNA-seq data 
capture a combination of these effects, as well as deterministic regulatory 
mechanisms [@Eling2019]. 
Moreover, these variability estimates can also be inflated by the technical 
noise that is typically observed in scRNA-seq data [@Brennecke2013]. This 
technical noise relates to systematic differences between cells that may be 
introduced by the data generating process (e.g., due to differences in dilution 
or sequencing depth) [@Vallejos2017]. 

<!--- Experimental strategies to tackle technical noise --->
Different strategies have been incorporated into scRNA-seq protocols to control 
or attenuate technical noise. 
For example, external RNA spike-in molecules (such as the set introduced by the 
External RNA Controls Consortium, ERCC [@Rna2005]) can be added to each cell’s 
lysate in a (theoretically) known fixed quantity.
Spike-ins can assist quality control steps [@McCarthy2017], data normalisation
[@Vallejos2017] and can be used to infer technical noise [@Brennecke2013].
Another strategy is to tag individual cDNA molecules using unique molecular 
identifiers (UMIs) before PCR amplification [@Islam2014]. 
Reads that contain the same UMI can be collapsed into a single molecule count,
attenuating technical variability associated to cell-to-cell differences
in amplification and sequencing depth (these technical biases are not fully 
removed unless sequencing to saturation [@Vallejos2017]). 
However, despite the benefits associated to the use of spike-ins and UMIs, 
these are not routinely available for all scRNA-seq protocols [@Haque2017]. 
In particular, spike-ins are of limited use in droplet-based protocols, 
as spike-ins can only be added to the reagent mixture in a known concentration,
and the exact quantity in each droplet necessarily remains unknown [@Bacher2016].

<!--- Introduce BASiCS --->
The Bioconductor package `r Biocpkg("BASiCS")` implements a Bayesian 
hierarchical framework that accounts for both technical and biological sources 
of noise in scRNA-seq datasets [@Vallejos2015; @Vallejos2016; @Eling2017]. The
model was initially motivated by supervised experimental designs in which 
experimental conditions correspond to groups of cells defined *a priori* (e.g. 
selected cell types obtained through FACS sorting). However, the approach can 
also be used in cases where the groups of cells of interest are computationally 
identified through clustering. In such cases, the model does not currently 
address issues associated to *post-selection inference*, where the same data is 
analysed twice: first to perform clustering and then to compare expression 
profiles between the clusters [@Lahnemann2020]; this is an inherent limitation 
of most differential expression tools.

`r Biocpkg("BASiCS")` jointly performs data normalisation, technical noise 
quantification and downstream analyses, whilst propagating statistical 
uncertainty across these steps. 
These features are implemented within a probabilistic model that builds upon a
negative binomial distribution, a widely used statistical model in the context 
of bulk and scRNA-seq experiments [@Love2014;@Svensson2020;@Townes2020;@Townes2019]. 
The negative binomial distribution is commonly used to model count data when
the observed variability differs from what can be captured by a simpler
Poisson model --- this is typically referred to as over-dispersion. 
<!--- emphasis, this is critical ---> 
Critically, `r Biocpkg("BASiCS")` enables the quantification of transcriptional
variability within a population of cells, while accounting for the overall 
mean-variance relationship that typically arises in scRNA-seq data [@Eling2018].
The latter enables the quantification of a gene-level *residual over-dispersion*
as a measure of transcriptional noise that is not confounded by differences in
gene expression. Furthermore, when available, `r Biocpkg("BASiCS")` can also 
leverage extrinsic spike-in molecules to aid data normalisation. 
A previous study has shown that BASiCS, when used as a generative
model, is well-tuned to capture and recapitulate the main properties of
typical scRNAseq data using posterior predictive checks [@Zappia2017].

<!--- Describes aims for the workflow and introduces BASiCS ---> 
This article complements existing scRNA-seq workflows based on the Bioconductor 
ecosystem (e.g. [@Lun2016; @Kim2019]), providing a detailed framework for 
transcriptional variability analyses using `r Biocpkg("BASiCS")`. 
We describe a step-by-step workflow that uses `r Biocpkg("scater")` 
[@McCarthy2017] and `r Biocpkg("scran")` [@Lun2016] to perform quality control 
(QC) as well as initial exploratory analyses.
Our analysis pipeline includes practical guidance to assess the convergence of 
the Markov Chain Monte Carlo (MCMC) algorithm that is used to infer model 
parameters in `r Biocpkg("BASiCS")`, as well as recommendations to interpret 
and post-process the model outputs. 
Finally, through a case study in the context of mouse immune cells, we illustrate
how `r Biocpkg("BASiCS")` can identify highly and lowly variable
genes within a cell population, as well as to compare expression profiles
between experimental conditions or cell types. 

All source code used to generate the results presented in this article is 
available [on Github](https://github.com/VallejosGroup/BASiCSWorkflow).
To ensure the 
reproducibility of this workflow, the analysis environment and all software 
dependencies are provided as a Docker image [@Boettiger2015]. The image 
can be obtained from
[Docker Hub](https://hub.docker.com/repository/docker/alanocallaghan/bocker).

# Methods {#methods}

This step-by-step scRNA-seq workflow is primarily based on the Bioconductor 
package ecosystem [@Amezquita2019]. 
A graphical overview is provided in Figure \@ref(fig:overview) 
and its main components are described below. 

```{r overview, out.width="0.7\\textwidth", fig.cap = "Graphical overview for the scRNA-seq analysis workflow described in this manuscript. Starting from a matrix of expression counts, we use the scater and scran Bioconductor packages to perform QC and initial exploratory analyses. To robustly quantify transcriptional heterogeneity within seemingly homogeneous cell populations, we apply the BASiCS Bioconductor package and  illustrate how BASiCS can be used to analyse a single or multiple pre-specified groups of cells.", echo=FALSE}
knitr::include_graphics("figure/Workflow-Overview.pdf", auto_pdf = TRUE)
```

Here, we load all the libraries that will be used throughout this workflow.
First, we load Bioconductor libraries included in Figure \@ref(fig:overview). 
We provide further explanation about their function in the following sections.

```{r}
library("SingleCellExperiment")
library("scater")
library("scran")
library("BASiCS")
```

Next, we load other general purpose libraries used for data visualisation (e.g.
`r CRANpkg("ggplot2")`), colour scales (e.g., `r CRANpkg("viridis")`,
and data manipulation (`r CRANpkg("tidyr")`). 

```{r}
# Data visualisation
library("ggplot2")
library("ggpointdensity")
library("patchwork")
library("ComplexHeatmap")
# colour scales
library("viridis")
library("circlize")
library("RColorBrewer")
# data manipulation
library("tidyr")
```

## Input data

We use the Bioconductor package `r Biocpkg("SingleCellExperiment")` to convert 
an input matrix of raw read counts (molecule counts for UMI-based protocols) 
into a `SingleCellExperiment` object that can also store its associated 
metadata, such as gene- and cell-specific information. 
Moreover, when available, the same object can also store read counts for 
spike-in molecules (see `help("altExp")`).
A major advantage of using `r Biocpkg("SingleCellExperiment")` is that it 
enables interoperability across a large number of Bioconductor packages [@Amezquita2019]. 

## QC and exploratory data analysis

A critical step in scRNA-seq analyses is QC, removing low quality samples that 
may distort downstream analyses. 
The [*OSCA*](https://bioconductor.org/books/release/OSCA/) online book provides 
an extensive overview on important aspects of how to perform QC of scRNA-seq 
data, including exploratory analyses [@Amezquita2019].
Here, we use QC diagnostics to identify and remove samples that correspond to 
broken cells, that are empty, or that contain multiple cells [@Ilicic2016]. 
We also remove lowly expressed genes that represent less reliable 
information.

We use the Bioconductor package `r Biocpkg("scater")` package [@McCarthy2017] to 
calculate QC metrics for each cell (e.g. total read-count) and gene 
(e.g. percentage of zeroes across all cells), respectively. 
We also use the visualisation tools implemented in the `r Biocpkg("scater")` to 
explore the input dataset and its associated QC diagnostic metrics. 
To perform further exploratory data analysis, we use the Bioconductor package 
`r Biocpkg("scran")` [@Lun2016].
The latter is used to perform *global scaling* normalisation, calculating 
cell-specific scaling factors that capture global differences in read-counts 
across cells (e.g. due to sequencing depth and PCR amplification) 
[@Lun2016pooling]. Note that, the use of `r Biocpkg("scran")` as a normalisation 
tool is not a requirement for `r Biocpkg("BASiCS")` (which uses raw read-counts 
as input). However, we recommend the use of `r Biocpkg("scran")` normalisation 
as part of the exploratory data analysis. For example, normalised read-counts 
can be used to visualise the data in a low-dimensional space (e.g. using 
principal component analysis). This can help to explore the structure of the 
data, e.g. to examine potential batch effects. 

## BASiCS - Bayesian Analysis of Single Cell Sequencing data

The `r Biocpkg("BASiCS")` Bioconductor package uses a Bayesian hierarchical 
model to simultaneously perform data normalisation, technical noise 
quantification and downstream analyses [@Vallejos2015BASiCS;@Vallejos2016;@Eling2018]
within a cell population or populations under study. In this context,
cell populations could correspond to groups set a priori by the experimental
design (e.g. naive or stimulated CD4+ T cells in [@Martinez-jimenez2017]),
or to groups of cells that were computationally identified through clustering.
Moreover, instead of modelling expression patterns separately for 
each gene, `r Biocpkg("BASiCS")` shares information between all genes to 
robustly quantify transcriptional variability. For example, as described by 
[@Eling2018], pooling information across genes with similar mean expression 
levels helps to obtain more reliable transcriptional variability estimates. The 
latter is particularly helpful for lowly expressed genes and sparse datasets, 
where less information is available.

A high-level overview for the model implemented in `r Biocpkg("BASiCS")` is
shown in Figure \@ref(fig:basics-schematic),
and a more detailed description is provided in the `ExtendedMethods` document
in the Zenodo repository for this manuscript (see [Data availability](#data-availability)).
Model parameters are designed to capture different 
aspects of scRNAseq data. First, *nuisance* parameters are used to 
capture technical artifacts. This includes cell-specific parameters (indexed by 
a $j$ subscript) used to perform global scaling normalisation (similar to the 
size factors in `r Biocpkg("scran")`) and global parameters designed to capture 
technical over-dispersion (this can also be interpreted as measurement error) 
that is shared by all genes. Note that, whilst `r Biocpkg("scran")` infers 
cell-specific size factors as a pre-processing step, `r Biocpkg("BASiCS")` uses 
an integrated approach wherein data normalisation and downstream analyses are 
performed simultaneously, thereby propagating statistical uncertainty. Secondly,
`r Biocpkg("BASiCS")` summarises expression patterns within the population of 
cells under study (e.g. a experimental condition or cell type) using two sets
of gene-specific parameters (indexed by a $i$ subscript). Mean parameters 
$\mu_i$ quantify overall expression for each gene $i$. In contrast, $\delta_i$ 
captures the excess of variability that is observed with respect to what would 
be expected in a homogeneous cell population, beyond technical noise. 
The latter is used as a proxy to quantify transcriptional variability.

To account for the strong relationship that is typically observed between
gene-specific mean expression and over-dispersion estimates, Eling *et al.* 
[@Eling2018] introduced a *joint prior* specification for these parameters.
This joint prior assumes that genes with similar mean expression ($\mu_i$) have
similar over-dispersion parameters $\delta_i$. Effectively, this shrinks
over-dispersion estimates towards a global trend that captures the relationship
between mean and over-dispersion (Figure \@ref(fig:basics-schematic)).
This improves posterior inference for over-dispersion parameters when the data
is less informative (e.g. small sample size, lowly expressed genes)
[@Eling2018]. This information-sharing approach is conceptually similar to
that performed by @Love2014 and others, where sparse data is pooled to obtain
more reliable estimates. The global trend is then used to derive gene-specific
*residual over-dispersion* parameters $\epsilon_i$ that are not confounded by
mean expression. Similar to the DM values implemented in `r Biocpkg("scran")`,
these are defined as deviations with respect to the overall trend.
These residual over-dispersion values can be used to quantify transcriptional
heterogeneity without any confounding with mean expression, though
it should be noted that they may also represent structural heterogeneity
resulting from heterogeneous cell types within the population being studied.

```{r basics-schematic, echo = FALSE, out.width="\\textwidth", fig.cap="A schematic representation of the features of the BASiCS model. BASiCS accounts for cell-to-cell and batch-to-batch variability using nuisance parameters. Accounting for cell-to-cell and batch-to-batch technical variability allows BASiCS to robustly perform gene-level inference of mean and variability. Furthermore, by accounting for the association between mean and variability in scRNAseq, BASiCS can also infer transcriptional noise using the residual variability parameter $\\epsilon_i$."}
knitr::include_graphics("figure/Workflow-Schematic.pdf", auto_pdf = TRUE)
```

Additionally, technical variation is quantified
using replication [@Carroll2005]. In the absence of true technical replicates,
we assume that population-level characteristics of the cells are replicated
using appropriate experimental design. This requires that cells from the same
population have been randomly allocated to different batches. Given appropriate
experimental design, `r Biocpkg("BASiCS")` assumes that biological effects
are shared across batches, while technical variation leads to spurious
differences between cells in different batches. Thus, `r Biocpkg("BASiCS")`
models cell-specific size factors $\nu_j$, and batch-specific technical
dispersion $\theta$.
It is this version of
the model that we focus on here, and that we recommend for most
users. Previous versions of the model are available within the package, but
are primarily useful for reproducibility purposes or for analysing
datasets that contain spike-in genes.


While several differential expression tools have been proposed for scRNA-seq 
data (e.g. [@Kharchenko2014; @Finak2015]), some evidence suggests that 
these do not generally outperform popular bulk RNA-seq tools [@Soneson2018]. 
Moreover, most of these methods are only designed to uncover changes in overall
expression, ignoring the more complex patterns that can arise at the single cell 
level [@Lahnemann2020]. 
Instead, `r Biocpkg("BASiCS")` embraces the high granularity of scRNA-seq data,
uncovering changes in cell-to-cell expression variability that are not 
confounded by differences in technical noise or in overall expression.

## Transcriptional variability analysis using `r Biocpkg("scran")` 

To quantify transcriptional variability, `r Biocpkg("scran")` can infer an 
overall trend between mean expression and the squared coefficient of 
variation (CV^2^) for each gene. 
Variability estimates that are not confounded by this 
trend are then obtained via the DM approach [@Kolodziejczyk2015cell]. 
For each gene, these are defined as the distance between CV$^2$ and a rolling 
median along the range of mean expression values.
DM estimates enable exploratory analyses of transcriptional 
variability, but a measure of statistical uncertainty is not readily available. 
As such, gene-specific downstream inference (such as differential variability 
testing) is precluded.

# Case study: analysis of presomitic and somitic mesoderm cells {#SM-analysis}

The development of droplet-based scRNA-seq [@Klein2015;@Macosko2015] 
lead to a large increase in the number of cells that can be profiled per 
experiment at a fixed cost.
With this, large-scale scRNA-seq datasets have been generated to study 
development across multiple time-points and capturing multiple tissues 
[@Ibarra-Soria2018;@Kernfeld2018]. 

To illustrate the use of `r Biocpkg("BASiCS")` in this context, we use 
droplet-based scRNAseq generated using the 10X microfluidics platform. An 
example using a dataset generated using a plate-based protocol including 
spike-ins is provided in the "AnalysisCD4T.Rmd" document of the
Zenodo repository for this manuscript (see [Data availability](#data-availability)).
In particular, we focus on the data generated by @Ibarra-Soria2018 which
contains expression profiles for 20,819 cells collected from E8.25 mouse 
embryos. These data comprise a highly heterogeneous population of cells with 
over 20 major cell types. The data is stored under the accession number 
E-MTAB-6153 on ArrayExpress and can be obtained from 
[here](https://www.ebi.ac.uk/biostudies/arrayexpress/studies/E-MTAB-6153#). 
The authors also provide additional information summarising the experimental
design (animal of source for each cell) and the cluster labels generated by
in their analysis. The code required to download these data is provided in 
the "DataPreparationBASiCSWorkflow.Rmd" document of
the Zenodo repository for this manuscript (see [Data availability](#data-availability)).

As explained above, when analysing a heterogeneous population of cells, a 
clustering analysis is required to identify the cell types that drive this 
heterogeneity prior to running `r Biocpkg("BASiCS")`. `r Biocpkg("BASiCS")` can 
then be used to quantify transcriptional variability within each cell type, as 
well as to characterise differences between cell types. Here, we use the cluster 
labels generated in the original publication. For illustration purposes, we 
focus on analysing presomitic and somitic mesoderm cells. This is an arbitrary 
choice and the steps described in this workflow could be applied to any other 
pair of cell types. 

## Preparing the input data for BASiCS

While quality control and pre-processing is an important topic in scRNAseq
analysis, we feel that it has been covered in detail in other articles.
For this analysis, we have performed quality control, pre-processing and 
exploratory data analysis in the "DataPreparationBASiCSWorkflow.Rmd" document of
the Zenodo repository for this manuscript (see [Data availability](#data-availability)).
The code presented there needs to be run before running the rest of this 
workflow. Here, we load the pre-processed data. This consists of two separate 
`SingleCellExperiment` data objects: one for presomitic and one for
somitic mesoderm cells. 

```{r SCE-load}
# Website were the files are located
chains_website <- "https://zenodo.org/record/10251224/files/"
# To avoid timeout issues as we are downloading large files
options(timeout = 1000)
# File download
## The code below uses `file.exists` to check if files were previously downloaded
## After download, files are then stored in an `rds` sub-folder 
if (!file.exists("rds/sce_sm.Rds")) {
  download.file(
    paste0(chains_website, "rds/sce_sm.Rds"),
    destfile = "rds/sce_sm.Rds"
  )
}
if (!file.exists("rds/sce_psm.Rds")) {
  download.file(
    paste0(chains_website, "rds/sce_psm.Rds"),
    destfile = "rds/sce_psm.Rds"
  )
}
# Pre-somitic mesoderm cells
sce_psm <- readRDS("rds/sce_psm.Rds")
# Somitic mesoderm cells
sce_sm <- readRDS("rds/sce_sm.Rds")
```

In brief, we have filtered the original data to select presomitic mesoderm (PSM) 
and somitic mesoderm (SM) cells, excluding a small cluster of cells which the 
authors identified a potential doublets (where multiple cells are captured in 
the same droplet). We also exclude cells with more 25,000 UMI-counts. The 
processed dataset contains `r ncol(sce_psm)` pre-somitic and `r ncol(sce_sm)` 
somitic mesoderm cells. 

Subsequently, we applied a gene filtering step to remove lowly expressed genes 
that were largely undetected through sequencing. In particular, we removed genes 
that are not detected in at least 20 cells across all cells. This is to ensure a 
reliable estimate of variability, and is roughly in line with the sample size 
requirements for the negative binomial distribution outlined in @Lloyd-Smith2007. 
We also applied a filter to consider protein-coding genes only. After applying 
these filters `r nrow(sce_psm)` will be included in our analysis. 

Since droplet-based scRNA-seq data are generated without including technical 
spike-in genes, BASiCS uses measurement error models to quantify technical 
variation through replication [@Carroll1998].
The different experimental batches used by @Ibarra-Soria2018 will be used for
this purpose. In the case of the somitic and pre-somoitic mesoderm cells, 
two mouse embryos have been used to generate the data. 
Cells isolated from the first embryo were split into two batches and processed 
independently. Cells from the second embryo were processed together as a single
batch. These three different groups of cells are used as batches. Numbers of 
cells per batch and cell type are displayed below.

```{r show-batches}
# Pre-somitic mesoderm cells
table(sce_psm$sample)
# Somitic mesoderm cells
table(sce_sm$sample)
```

To enable BASiCS to use this information, this should be stored as `BatchInfo`
within the cell-level metadata. 

```{r BASiCS-droplet-data}
# Pre-somitic mesoderm cells
colData(sce_psm)$BatchInfo  <- colData(sce_psm)$sample
# Somitic mesoderm cells
colData(sce_sm)$BatchInfo  <- colData(sce_sm)$sample
```

## Parameter estimation using BASiCS

Parameter estimation is implemented in the `BASiCS_MCMC` function using an 
adaptive Metropolis within Gibbs algorithm (see section 3 in [@Roberts2009]).
The primary inputs for `BASiCS_MCMC` correspond to:

- `Data`: a `SingleCellExperiment` object created as described in the 
  previous sections. 
- `N`: the total number of MCMC iterations.
- `Thin`: thining period for output storage
  (only the `Thin`-th MCMC draw is stored).
- `Burn`: the initial number of MCMC iterations to be discarded.
- `Regression`: if `TRUE` a joint prior is assigned to $\mu_i$ and $\delta_i$ 
  [@Eling2018], and residual over-dispersion values $\epsilon_i$ are inferred. 
  Alternatively, independent log-normal priors are assigned to $\mu_i$ and 
  $\delta_i$ [@Vallejos2016]. 
- `WithSpikes`: if `TRUE` information from spike-in molecules is used to aid
  data normalisation and to quantify technical noise. 
- `PriorParam`: Defines the prior hyper-parameters to be used by `r Biocpkg("BASiCS")`.

As a default, we recommend to use `Regression = TRUE`, as the joint prior 
introduced by [@Eling2018] leads to more stable estimation,
particularly for small sample sizes and lowly expressed genes. 
This approach also enables users to obtain a measure of
transcriptional variability that is not confounded by mean expression. 
Advanced users may use the `BASiCS_PriorParam` function
to adjust prior hyperparameters. If `PriorMu = "EmpiricalBayes"`, $\mu_i$'s are assigned a 
log-normal prior with gene-specific location hyper-parameters defined via an 
empirical Bayes framework. Alternatively, if `PriorMu = "default"`, location 
hyper-parameters are set to be equal 0 for all genes. We recommend that users
use the defaults for `PriorParam` for most applications.
We recommend to use `PriorMu = "EmpiricalBayes"` as we have observed 
that an empirical Bayes framework [@Casella1985]
improves estimation performance for sparser datasets. 
Extra parameters can be used to store the output 
(`StoreChains`, `StoreDir`, `RunName`) and to monitor the progress of the 
algorithm (`PrintProgress`).


Here, we run the MCMC sampler separately for somitic and pre-somitic mesoderm
cells. We use 20,000 iterations (`N`), discarding the initial 10,000 iterations
as burn-in, (`Burn`), and saving parameter values only once in each 10
iterations (`Thin`).
We recommend this setting for large datasets, as it generally produces reliable
and reproducible results. However, specific datasets may require a larger number
of iterations to achieve convergence. Practical guidance about MCMC convergence 
diagnostics is provided in the next section.

```{r MCMC-run, eval = FALSE}
## MCMC results may vary slightly due to pseudorandom number generation.
## We fix a random seed for exact reproducibility,
## but this is not strictly required
set.seed(42)
chain_sm <- BASiCS_MCMC(
  Data = sce_sm,
  N = 30000,
  Thin = 15,
  Burn = 15000,
  Regression = TRUE,
  WithSpikes = FALSE,
  PriorParam = BASiCS_PriorParam(sce_sm, PriorMu = "EmpiricalBayes"),
  Threads = 4,
  StoreChains = TRUE,
  StoreDir = "rds/",
  RunName = "SM"
)
set.seed(43)
chain_psm <- BASiCS_MCMC(
  Data = sce_psm,
  N = 30000,
  Thin = 15,
  Burn = 15000,
  Regression = TRUE,
  WithSpikes = FALSE,
  PriorParam = BASiCS_PriorParam(sce_psm, PriorMu = "EmpiricalBayes"),
  Threads = 4,
  StoreChains = TRUE,
  StoreDir = "rds/",
  RunName = "PSM"
)
```

This first of these chains takes 140 minutes to complete on a 3.4 GHz Intel
Core i7 4770k procesor with 32GB RAM, while the second takes 159 minutes.
For convenience, these can be obtained online at 
[https://doi.org/10.5281/zenodo.10251224](https://doi.org/10.5281/zenodo.10251224).

```{r download-chain-SM}
# Website were the files are located
chains_website <- "https://zenodo.org/record/10251224/files/"
# To avoid timeout issues as we are downloading large files
options(timeout = 1000)
# File download
## The code below uses `file.exists` to check if files were previously downloaded
## After download, files are then stored in an `rds` sub-folder 
if (!file.exists("rds/chain_sm.Rds")) {
  download.file(
    paste0(chains_website, "rds/chain_sm.Rds"),
    destfile = "rds/chain_sm.Rds"
  )
}
if (!file.exists("rds/chain_psm.Rds")) {
  download.file(
    paste0(chains_website, "rds/chain_psm.Rds"),
    destfile = "rds/chain_psm.Rds"
  )
}

## Loads files after download
chain_sm <- readRDS("rds/chain_sm.Rds")
chain_psm <- readRDS("rds/chain_psm.Rds")
```

The output from `BASiCS_MCMC` is a `BASiCS_Chain` object that contains the 
draws associated to all model parameters. Given that 
`N` = 30,000, `Thin` = 15 and `Burn` = 15,000, the object contains 1,000 
draws for each parameter.
These can be accessed using the `displayChainBASiCS` function.
For example, the following code displays the first 2 MCMC draws for mean
expression parameters $\mu_i$ associated to the first 3 genes.

```{r displayChainBASiCS}
displayChainBASiCS(chain_sm, Param = "mu")[1:2, 1:3]
```


## MCMC convergence diagnostics

Before interpreting the estimates generated by `r Biocpkg("BASiCS")`, 
it is important to check the performance of the MCMC algorithm used.
This algorithm used to characterise the posterior distribution is stochastic
by nature, and as such its performance may vary even when used on the same
dataset using the same settings. We perform quality control of the MCMC
algorithm in order to ensure that the results we observe are the result of
properties of the underlying data, rather than being an artefact of the
stochastic algorithm used to characterise the data.

As part of performing quality control of MCMC performance,
it is critical to assess the convergence of the MCMC algorithm, i.e. whether
the MCMC reached its stationary distribution.
If convergence has been achieved, the trace for each parameter should not evolve
significantly across iterations, as MCMC draws are expected to be stochastic 
fluctuations around a horizontal trend once the sampler has converged to its
stationary distribution.
It is not possible to prove convergence, but multiple graphical and
quantitative convergence diagnostics have been proposed to assess the lack
of convergence (e.g. [@CowlesCarlin1996; @BrooksGelman1998]). 
Some advocate the use of multiple MCMC chains using different starting values in
order to ensure that the algorithm consistently converges to the same 
distribution and to avoid convergence to local modes.
For `r Biocpkg("BASiCS")`, we have observed that using informed starting values 
(e.g. based on `r Biocpkg("scran")` normalisation factors) and a sufficiently 
large value for `N` and `Burn` generally leads to largely consistent estimates 
across multiple MCMC runs.
Hence, the focus of this section is to evaluate quantitative measures of 
convergence (e.g. [@Geweke1995]) based on a single MCMC chain.

For illustration purposes, here we focus on the chains generated for the 
somatic mesoderm cells. However, it is important to analyse convergence for 
all MCMC chains; we have omitted this analysis for pre-somitic 
mesoderm cells in the present manuscript for brevity, but they can be viewed
in the data preparation document shown in the Zenodo repository for
this manuscript (see [Data availability](#data-availability)).

Traceplots can be used to visually assess the history of MCMC iterations 
for a specific parameter (e.g. Figure \@ref(fig:convergence-SM), left panel). 
As mentioned above, significant departures from a horizontal trend suggest a
lack of convergence.
As illustrated in Figure \@ref(fig:convergence-SM), histograms can also be 
used to display the marginal distribution for each parameter. 
For `r Biocpkg("BASiCS")`, users should expect these to follow a unimodal 
distribution. Failure to satisfy these graphical diagnostics suggest that `N` 
and `Burn` must be increased. 
Alternatively, more stringent quality control could be applied to the input 
data, as we have observed that genes with very low counts often suffer from 
slow convergence.

```{r convergence-SM, fig.cap="Trace plot, marginal histogram, and autocorrelation function of the posterior distribution of the mean expression parameter for a gene in somitic mesoderm cells following MCMC sampling. Trace plots should explore the posterior without getting stuck in one location or drifting over time towards a region of higher density. High autocorrelation indicates that the number of effective independent samples is low. It is good practice to perform these visualisation for many different parameters; here we only show one."}
plot(chain_sm, Param = "mu", Gene = 1)
```

As `r Biocpkg("BASiCS")` infers thousands of parameters, it is 
impractical to assess these diagnostics separately for each parameter. 
Thus, it is helpful to use numerical diagnostics that can be applied to multiple
parameters simultaneously. 
The `BASiCS_DiagPlot()` can be used to visualise such diagnostic metrics. 
Here, we illustrate usage for two such metrics focusing on the MCMC
chain that was obtained for the somitic mesoderm cells (similar results were 
obtained for pre-somitic mesoderm cells in the
"DataPreparationBASiCSWorkflow.Rmd" document of
the Zenodo repository (see [Data availability](#data-availability)). 
First, we focus on the diagnostic criterion proposed by Geweke [@Geweke1995].
The latter compares the average of draws obtained during the initial (10\% after 
burn in, by default) and the final part of the chain (50\% by default)
by calculating $Z$-scores of the relative difference between the two sets 
of draws.
Large absolute $Z$-scores suggest that the algorithm has not converged (as a
rule of thumb, a threshold at $|Z| < 3$ is often applied). 
For the somitic and pre-somitic mesoderm datasets, most Z-scores associated to mean
expression parameters $\mu_i$ were small in absolute value 
(see Figure \@ref(fig:mcmc-diag)A), suggesting that the algorithm has 
largely converged.


As well as assessing MCMC convergence, it is important to ensure that the MCMC 
algorithm has efficiently explored the parameter space. 
For example, the autocorrelation function (e.g. Figure 
\@ref(fig:convergence-SM), right panel) quantifies the 
correlation between the chain and its lagged versions. 
Strong autocorrelation indicates that neighbouring MCMC samples are highly 
dependent and suggest poor sampling efficiency. 
The latter may indicate that the MCMC draws do not contain 
sufficient information to produce accurate posterior estimates.
In other words, highly correlated MCMC samplers require more samples to produce
the same level of Monte Carlo error for an estimate 
(defined as the variance of a Monte Carlo estimate across repetitions [@Koehler2009]). 

The effective sample size (ESS) is a related measure which represents a proxy 
for the number of independent draws generated by the MCMC sampler [@Gelman2014].
The latter is defined as:
$$
  \mbox{ESS} = \frac{N_{tot}}{1 + 2\sum_{k=1}^\infty \rho(k)},
$$
where $N_{tot}$ represents the total number of MCMC draws (after burn-in and
thining) and $\rho(k)$ is the autocorrelation at lag $k$.
ESS estimates associated to mean expression parameters for the somitic mesoderm
cells are displayed in Figure \@ref(fig:mcmc-diag)B.
Whilst ESS is around 1,000 ($N_{tot}$ in this case) for most genes, we observe 
low ESS values for a small proportion of genes (primarily lowly expressed genes 
whose expression was only captured in a small number of cells).
As described later in this manuscript, `BASiCS_TestDE` automatically excludes 
genes with low ESS during differential expression testing (by default
a threshold at ESS $< 100$ is applied). However, if a large number of genes have
large Geweke diagnostic values or low effective sample sizes in a certain 
dataset, then caution should be applied when interpreting the results of the 
model. These issues can often be addressed by more stringent 
filtering of genes and cells before performing inference.

```{r mcmc-diag, fig.cap="MCMC diagnostics for gene-specific mean expression parameters; somitic mesoderm cells. A: Geweke Z-score for mean expression parameters is plotted against mean expression estimates. Dashed lines represent absolute Z-scores of 3, outside of which we advise caution when interpreting results. B: Effective sample size (ESS) is plotted against mean expression estimates. A dashed line shows a threshold of 100, below which we advise caution when interpreting results."}
diag_p1 <- BASiCS_DiagPlot(chain_sm, Param = "mu", Measure = "geweke") +
  theme(legend.position = "bottom")
diag_p2 <- BASiCS_DiagPlot(chain_sm, Param = "mu", Measure = "ess") +
  theme(legend.position = "bottom")
diag_p1 + diag_p2 + plot_annotation(tag_levels = "A")
```

## Quantifying transcriptional variability using BASiCS

Studying gene-level transcriptional variability can provide insights about the 
regulation of gene expression, and how it relates to the properties of genomic 
features (e.g. CpG island composition [@Morgan2018]), transcriptional dynamics 
[@Antolovic2017] and aging [@Martinez-jimenez2017], among others. 
The squared coefficient of variation (CV^2^) is widely used as a proxy for 
transcriptional variability. 
For example, we can obtain CV^2^ estimates for each gene using 
`r Biocpkg("scran")` normalised counts as input. 
In contrast, `r Biocpkg("BASiCS")` infers transcriptional variability using 
gene-specific over-dispersion parameters $\delta_i$ (see *Methods*).
Figure \@ref(fig:dispersion-vs-mean) compares these approaches, focusing on 
somitic mesoderm cells (repeating this analysis for pre-somitic mesoderm cells 
led to similar results). 
For each of the panels in  Figure \@ref(fig:dispersion-vs-mean),
we use the R package `r CRANpkg("ggpointdensity")` to visualise 
the local density of genes along the axes.
Note that the `BASiCS_ShowFit` 
function can be used to generate Figure \@ref(fig:dispersion-vs-mean)C, but we 
generated the plot manually to demonstrate how users can extract this 
information from a `BASiCS_MCMC` object, and for visual consistency with the 
other panels.

```{r dispersion-vs-mean, fig.width=6, fig.height=6, fig.cap="Comparison of gene-specific transcriptional variability estimates and mean expression estimates obtained for each gene using BASiCS and scran. For this analysis, we exclude genes that are not expressed in at least 2 cells. BASiCS estimates for each gene are defined by the posterior median of the associated parameter. scran estimates for each gene are derived after applying the pooling normalisation strategy proposed by Lun et al. Points are coloured according to the local density of genes along the x- and y- axis. A: scran squared CV estimates versus BASiCS estimates for over-dispersion parameters. B: scran estimates for mean expression and the squared CV. C: BASiCS estimates for mean expression and over-dispersion parameters. D: BASiCS estimates for residual over-dispersion parameters versus DM values estimated by scran. E: scran estimates for mean expression and DM values. F: BASiCS estimates for mean expression and residual over-dispersion parameters. Dashed red lines in panels A and D represent the line given by x=y. "}

## Get BASiCS point estimates for mean and variability - SM cells
summary_sm <- Summary(chain_sm)
parameter_df <- data.frame(
  mean = displaySummaryBASiCS(summary_sm, Param = "mu")[, 1],
  overdisp = displaySummaryBASiCS(summary_sm, Param = "delta")[, 1],
  residualoverdisp = displaySummaryBASiCS(summary_sm, Param = "epsilon")[, 1]
)

## Calculate scran size factors - SM cells
sce_sm <- computeSumFactors(sce_sm)
## Get scran estimates for mean and variability - SM cells
sce_sm <- logNormCounts(sce_sm, log = FALSE)
parameter_df$mean_scran <- rowMeans(assay(sce_sm, "normcounts"))
parameter_df$cv2_scran <- rowVars(assay(sce_sm, "normcounts")) /
  parameter_df$mean_scran^2
parameter_df$DM <- DM(
  mean = parameter_df$mean_scran,
  cv2 = parameter_df$cv2_scran
)

## Remove genes without counts in > 2 cells - BASiCS estimates not provided
sel_genes <- !is.na(parameter_df$residualoverdisp) 

## The plots below are generated using ggplot2
## We use `plot_params` to specify the format of 
## such plots whilst avoiding code duplication
## The plots can be further customized using ggplot2 principles
## For example, we modify x- and y-axis limits in some of the plots below
plot_params <- list(
  geom_pointdensity(size = 0.6, na.rm = TRUE),
  scale_colour_viridis(name = "Density"),
  theme(
    text = element_text(size = rel(3)),
    legend.position = "bottom",
    legend.text = element_text(angle = 90, size = 8, hjust = 0.5, vjust = 0.5),
    legend.key.size =  unit(0.018, "npc")
  )
)

g1 <- ggplot(parameter_df[sel_genes,], aes(log10(cv2_scran), log10(overdisp))) +
  xlim(-3, 3) + ylim(-3, 3) +
  geom_abline(
    slope = 1,
    intercept = 0,
    colour = "firebrick",
    linetype = "dashed"
  )
  
g2 <- ggplot(parameter_df[sel_genes,]) +
  aes(log10(mean_scran), log10(cv2_scran)) +
  xlim(-3, 2.2) + ylim(-3, 3) 

g3 <- ggplot(parameter_df[sel_genes,], aes(log10(mean), log10(overdisp))) +
  xlim(-3, 2.2) + ylim(-3, 3) 

g4 <- ggplot(parameter_df[sel_genes,], aes(DM, residualoverdisp)) +
  geom_abline(
    slope = 1,
    intercept = 0,
    colour = "firebrick",
    linetype = "dashed"
  )

g5 <- ggplot(parameter_df[sel_genes,], aes(log10(mean_scran), DM)) 

g6 <- ggplot(parameter_df[sel_genes,], aes(log10(mean), residualoverdisp))

((g1 + g2 + g3) * plot_params) / ((g4 + g5 + g6) * plot_params)  +
  plot_annotation(tag_levels = "A") & theme(plot.tag = element_text(size = 15))
```


As seen in Figure \@ref(fig:dispersion-vs-mean)A, CV^2^ and posterior estimates 
for $\delta_i$ are highly correlated. 
Moreover, both variability metrics are confounded by differences in mean 
expression, i.e. highly expressed genes tend to exhibit lower variability
(Figures \@ref(fig:dispersion-vs-mean)B-C). 
We note, however, that `r Biocpkg("scran")` infers a more narrow distribution of
transcriptional variability estimates for any given mean expression level.
To remove the confounding observed in Figures \@ref(fig:dispersion-vs-mean)B-C, 
`r Biocpkg("scran")` and `r Biocpkg("BASiCS")` derive *residual variability* 
estimates as deviations with respect to an global mean-variability trend 
(see *Methods*).
These are derived using the DM approach [@Kolodziejczyk2015cell] and the
residual over-dispersion parameters $\epsilon_i$ defined by [@Eling2018],
respectively.
For the somitic mesoderm cell data, both approaches led to correlated 
estimates (Figure \@ref(fig:dispersion-vs-mean)D, Pearson's correlation equal 
to `r with(parameter_df, cor(DM, residualoverdisp))`) but the residual 
variability estimates generated for these data are less consistent than what we 
have observed in the context of less sparse data (see the "AnalysisCD4T.Rmd"
document in the Zenodo repository for this manuscript (see [Data availability](#data-availability)). 
For the majority of genes 
(`r with(parameter_df[sel_genes,], sum(sign(DM) == sign(residualoverdisp)))`
out of `r sum(sel_genes)`
captured in at least 3 
cells) both approaches led some broadly similar results, inferring the same 
sign for their residual variability estimates.

Finally, as expected, we observe that
neither DM values nor posterior estimates for $\epsilon_i$ are 
associated with mean expression (Figure \@ref(fig:dispersion-vs-mean)E-F).
However, unlike the DM method, the integrated approach implemented in 
`r Biocpkg("BASiCS")` provides a direct measure of statistical uncertainty for 
these estimates via posterior variance. 


## HVG/LVG detection using BASiCS

In `r Biocpkg("BASiCS")`, the functions `BASiCS_DetectHVG` and 
`BASiCS_DetectLVG` can be used to identify genes with substantially high (HVG) 
or low (LVG) transcriptional variability within a population of cells. 
If the input `BASiCS_Chain` object was generated by `BASiCS_MCMC` with 
`Regression = TRUE` (recommended setting), this analysis is based on the 
posterior distribution obtained for gene-specific residual over-dispersion 
parameters $\epsilon_i$ (alternatively, the approach introduced by 
[@Vallejos2015] can be used). 
HVGs are marked as those for which $\epsilon_i$ exceeds a pre-defined threshold 
with high probability, where the probability cut-off is chosen to match a given 
expected false discovery rate (EFDR; by default the target EFDR is set to 10\%) 
[@Newton2004].
The expected false discovery rate we use is conceptually similar to false
discovery rate control procedures in frequentist statistics, such
as the approach of @Benjamini1995, aiming to control the proportion of false
discoveries produced by the procedure.
A similar approach is implemented for LVG detection, but based 
on whether $\epsilon_i$ is below a pre-specified threshold `EpsilonThreshold`. 
For example, if the threshold for $\epsilon_i$ is equal to $\log(2)$ 
(the default), HVGs would 
be those genes for which the over-dispersion is estimated to be at 
least $2$ times higher than would be expected given the inferred mean 
expression level, while LVGs would be those genes for which the residual 
over-dispersion is at least $2$ time lower than would be expected on the same 
basis. In some circumstances, it may be of interest to rank genes and 
to select the those with the highest or the lowest residual over-dispersion,
which can be performed using the `PercentileThreshold` parameter; see 
`help("BASiCS_DetectVG")` for more details.

```{r det-vg-SM, message=TRUE, fig.cap="HVG and LVG detection using BASiCS. For each gene, BASiCS posterior estimates (posterior medians) associated to mean expression and residual over-dispersion parameters are plotted. Genes are coloured according to HVG/LVG status. Genes that are not expressed in at least 2 cells are excluded."}
## Highly variable genes
hvg <- BASiCS_DetectHVG(chain_sm, EpsilonThreshold = log(2))
## Lowly variable genes
lvg <- BASiCS_DetectLVG(chain_sm, EpsilonThreshold = -log(2))
```

For subsequent visualisation and data exploration, here we merge the output
tables generated above and store these as part of the gene-level metadata
in the `sce_sm` object.

```{r metadata-VG-SM}
## Merge HVG and LVG tables in a single data frame
vg_table <- merge(
  as.data.frame(lvg, Filter = FALSE),
  as.data.frame(hvg, Filter = FALSE),
  by = c("GeneName", "GeneIndex", "Mu", "Delta", "Epsilon"),
  suffixes = c("LVG", "HVG")
)
## Mark genes as highly variable, lowly variable, or not either.
vg_table$VG <- "Not HVG or LVG"
vg_table$VG[vg_table$HVG] <- "HVG"
vg_table$VG[vg_table$LVG] <- "LVG"
## Store as gene-level metadata
rowData(sce_sm) <- merge(
  rowData(sce_sm), vg_table, 
  by.x = "ensembl_gene_id", by.y = "GeneName",
  sort = FALSE
)
```

Performing HVG and LVG detection involves identifying a posterior probability
threshold that matches a target EFDR. We perform this choice using a 
grid search in which EFDR is calculated for a range of different probability 
thresholds between 0.5 and 1. As seen in Figure \@ref(fig:efdr-plot-vg-SM),
for the somitic mesoderm cells, this leads to a threshold of 
`r hvg@ProbThreshold` and `r lvg@ProbThreshold` for HVG and LVG detection, 
respectively (the value of these thresholds can be extracted from the 
`ProbThreshold` slot in the `hvg` and `lvg` objects, e.g. using 
`hvg@ProbThreshold`). 
For some datasets, the grid search may fail to identify a probability threshold
that matches the target EFDR. In such cases, the default minimum probability 
threshold of $\frac{2}{3}$ is used. 
This default threshold value can be changed using the `ProbThreshold` to
`BASiCS_DetectHVG` and `BASiCS_DetectLVG`.

```{r efdr-plot-vg-SM, fig.cap = "Lines representing the EFDR and EFNR for a range of posterior probability thresholds for the HVG (A) and LVG (B) detection tasks. The target EFDR is indicated by dashed horizontal lines, while the chosen posterior probability threshold is indicated by a vertical dashed line."}
p1 <- BASiCS_PlotVG(hvg, Plot = "Grid")
p2 <- BASiCS_PlotVG(lvg, Plot = "Grid") 

p1 + p2 +
  plot_annotation(tag_levels = "A") +
  plot_layout(guides = "collect") &
  theme(legend.position = "bottom")
```

```{r plot-vg-SM, fig.cap="HVG and LVG detection using BASiCS. For each gene, BASiCS posterior estimates (posterior medians) associated to mean expression and residual over-dispersion parameters are plotted. Genes are coloured according to HVG/LVG status. Genes that are not expressed in at least 2 cells are excluded."}
plotRowData(sce_sm, x = "Mu", y = "Epsilon", colour_by = "VG") + 
  scale_x_log10() + 
  geom_hline(yintercept = c(-log(2), 0, log(2)), lty = 2) +
  labs(
    x = "BASiCS means (log10)",
    y = "BASiCS residual\nover-dispersion"
  ) 
```

For the somitic mesoderm cells data, we obtained 
*`r sum(hvg@Table$HVG, na.rm=TRUE)`* HVG and *`r sum(lvg@Table$LVG, na.rm=TRUE)`* LVG.
As shown in Figure \@ref(fig:plot-vg-SM), these genes are distributed across 
a wide range of mean expression values. 
As an illustration, Figure \@ref(fig:plot-example-vg-SM) shows the 
distribution of normalised expression values for selected HVG and LVG, focusing 
on examples with similar mean expression levels.
As expected, HVG tend to exhibit a wider and potentially bimodal
distribution (Figure \@ref(fig:plot-example-vg-SM)A).
Instead, LVG tend to have more narrow and unimodal distributions 
(Figure \@ref(fig:plot-example-vg-SM)B).

```{r plot-example-vg-SM, fig.cap = "BASiCS denoised counts for example HVG (A) and LVG (B) with similar overall levels of expression."}
## Obtain normalised expression values and store them as an assay in sce_sm
assay(sce_sm, "BASiCS_norm") <- BASiCS_DenoisedCounts(
  sce_sm, chain_sm, WithSpikes = FALSE
)

## Select HVG/LVG genes with similar mean expression values
low_exp <- 20
up_exp <- 40
is_mid_exp <- rowData(sce_sm)$Mu > low_exp & rowData(sce_sm)$Mu < up_exp
is_hvg <- is_mid_exp & rowData(sce_sm)$HVG
is_lvg <- is_mid_exp & rowData(sce_sm)$LVG

## Amongst those, order by epsilon and select top 5 HVG and LVG 
top_hvg <- order(rowData(sce_sm)$Epsilon[is_hvg], decreasing = TRUE)[1:5]
top_lvg <- order(rowData(sce_sm)$Epsilon[is_lvg], decreasing = FALSE)[1:5]

## Generate violin plots for the selected HVG and LVG
## using scater::plotExpression
plot_hvg <- plotExpression(sce_sm, 
    rowData(sce_sm)$external_gene_name[is_hvg][top_hvg],
    exprs_values = "BASiCS_norm",
    swap_rownames = "external_gene_name",
    feature_colours = FALSE,
    point_fun = function(...) list() # disables drawing points
  ) +
  ylim(0, 90) 

plot_lvg <- plotExpression(
    sce_sm,
    rowData(sce_sm)$external_gene_name[is_lvg][top_lvg], 
    exprs_values = "BASiCS_norm",
    swap_rownames = "external_gene_name",
    feature_colours = FALSE,
    point_fun = function(...) list() # disables drawing points
  ) +
  ylim(0, 90) 

plot_hvg + plot_lvg + plot_annotation(tag_levels = "A")
```

## Differential mean and variability testing using BASiCS

This section highlights the use of `r Biocpkg("BASiCS")` to perform differential 
expression tests for mean and variability between different pre-specified 
populations of cells and experimental conditions. 
Here, we compare the somitic mesoderm cells, analysed in the previous section,
to pre-somitic mesoderm cells analysed in the same study.
Differential expression testing is performed via the
`BASiCS_TestDE` function. The main input parameters are

- `Chain1` and `Chain2`: two `BASiCS_Chain` objects created via the `BASiCS_MCMC`
function. Each object corresponds to a different pre-specified group of cells
- `EpsilonM` and `EpsilonR`: introduce a minimum effect size (in a
$\log_2$ fold change scale) for the detection of changes in mean or residual 
over-dispersion, respectively. 
This enables us to discard small expression changes that are less biologically
meaningful. By default, we set these thresholds to be equivalent to a 50\% 
change between the groups. 
However, different thresholds may be required depending on the context.
For example, if most genes show strong differences in mean expression, it can be 
beneficial to increase the value of `EpsilonM` to focus on strong changes in 
mean expression.
- `EFDR_M` and `EFDR_R`: define the target EFDR to calibrate the 
decision rule associated to changes in mean or residual over-dispersion, 
respectively. Default: 10\%.
- `MinESS`: genes with ESS values below this threshold will be excluded from the
differential expression tests.
This is used to increase the robustness of the results, excludes genes for which 
the sampler explored the parameter space less efficiently (see 
*MCMC diagnostics* Section). Default: `MinESS` = 100. 

```{r mean-expression-testing}
## Perform differential testing
test_de <- BASiCS_TestDE(
  Chain1 = chain_sm,
  Chain2 = chain_psm,
  GroupLabel1 = "SM",
  GroupLabel2 = "PSM",
  EFDR_M = 0.1,
  EFDR_R = 0.1,
  MinESS = 100,
  Plot = FALSE,
  PlotOffset = FALSE
)
table_de_mean <- as.data.frame(
  test_de,
  Parameter = "Mean",
  Filter = FALSE
)
table_de_resdisp <- as.data.frame(
  test_de,
  Parameter = "ResDisp",
  Filter = FALSE
)
## combine results of differential mean and residual over-dispersion tests
table_de <- merge(table_de_mean, table_de_resdisp)
## Merge with gene-level metadata to obtain gene names used in figures
table_de <- merge(
  table_de, 
  rowData(sce_sm)[, c("ensembl_gene_id", "external_gene_name")],
  by.x = "GeneName", by.y = "ensembl_gene_id"
)
# convert to standard data.frame, not S4 DFrame, for ggplot2
table_de <- as.data.frame(table_de)
```

After running the test, it is important to visualise the results to facilitate
interpretation and to identify systematic patterns among differentially
expressed genes. It may also be useful to perform functional enrichment 
analysis to identify biologically meaningful patterns among these genes. 
For example, this could be performed using the Bioconductor package
`r Biocpkg("goseq")` [@Young2010].
We do not perform this here, but a relevant workflow is described by
@Maksimovic2020.

We first focus on the differential mean expression test. MA-plots
(log fold change M versus mean average A) and volcano plots (posterior 
probability versus log fold change) are popular graphical summaries in this
context, and are presented in Figure \@ref(fig:visualise-DE-mean-plot). These
can be useful in ensuring that suitable magnitude and confidence thresholds have
been chosen. In this instance, it is clear that a large number of genes are
differentially expressed between the two conditions, and the selected
probability threshold is suitable.

```{r visualise-DE-mean-plot, fig.height = 8, fig.width = 6, fig.cap = "Upper panel presents the MA plot associated to the differential mean expression test between somitic and pre-somitic cells. Log-fold changes of average expression in somitic cells relative to pre-somitic cells are plotted against average expression estimates combined across both groups of cells. Bottom panel presents the volcano plot associated to the same test. Log-fold changes of average expression in somitic cells relative to pre-somitic mesoderm cells are plotted against their associated tail posterior probabilities. Colour indicates the differential expression status for each gene, including a label to identify genes that were excluded from differential expression test due to low ESS."}
p1 <- BASiCS_PlotDE(test_de, Parameters = "Mean", Plots = "MA")
p2 <- BASiCS_PlotDE(test_de, Parameters = "Mean", Plots = "Volcano",
  TransLogit = TRUE) ## logit-transforms the Y-axis, which can be clearer.
p1 / p2
```

When interpreting the results of differential expression tests, it is useful
to visualise expression patterns for differentially expressed genes in order to 
appraise the significance of the results and guide interpretation. 
For this purpose, we obtain normalised expression values for each group of 
cells after correcting for global changes in overall expression between the 
groups, i.e. a global offset that leads to uniformly higher expression across 
genes within one group of cells (such step is also internally done by 
`BASiCS_TestDE`). Among other causes, the latter can be due to overall 
differences in mRNA content between the groups. 

```{r offset-denoised}
## Calculate a global offset between groups, using the SM population as reference
offset <- BASiCS_CorrectOffset(chain_psm, chain_sm)
offset$Offset
## Get offset corrected chain
chain_psm <- offset$Chain
## Obtain normalised counts within for the PSM population
## Normalised counts were for the SM population were already obtained
## in the HVG/LVG analysis section
## Obtain normalised expression values and store them as an assay in sce_sm
assay(sce_psm, "BASiCS_norm") <- BASiCS_DenoisedCounts(
  sce_psm, chain_psm, 
  WithSpikes = FALSE
)
denoised_counts <- SingleCellExperiment(
  assays = list(
    BASiCS_norm = cbind(
      assay(sce_psm, "BASiCS_norm"),
      assay(sce_sm, "BASiCS_norm")
    )
  ),
  colData = data.frame(
    Celltype = c(
      rep("presomitic", ncol(sce_psm)),
      rep("somitic", ncol(sce_sm))
    )
  )
)
```

To visualise expression patterns for multiple genes at once, we use the 
Bioconductor package `r Biocpkg("ComplexHeatmap")` [@Gu2016] package, grouping
genes according to the result of the differential mean expression test 
(i.e. up-regulated in somitic/pre-somitic cells or non differentially expressed;
see Figures \@ref(fig:heatmap-diffexp1), \@ref(fig:heatmap-diffexp2) and \@ref(fig:heatmap-diffexp3)). 
For example, among the non DE group, we observe several genes encoding
ribosomal proteins (e.g. *Rps14*). 
Genes in this family have been previously observed to have stable expression
across a wide range of scRNAseq datasets in mouse and human [@Lin2019].
Such visualisations may aid in the interpretation of such stable or 
"housekeeping" genes, as well as genes which are up- or down-regulated in each
population.


To do this in each dataset, we create a few small utility functions,
starting with a function to select the genes with highest probability
of being differentially expressed, or the genes with smallest probability of
being differentially expressed:

```{r select-top-n}
select_top_n <- function(table, condition, n=15, decreasing=TRUE) {
  ind_condition <- table$ResultDiffMean == condition
  table <- table[ind_condition, ]
  ind_diff <- order(table$ProbDiffMean, decreasing = decreasing)[1:n]
  return(table$GeneName[ind_diff])
}
```

Next, we define a function that uses this gene selection to subset the
normalised counts for each population:

```{r use-select}
use_select <- function(sce, psm, sce_sm, select) {
  counts_psm <- assay(sce_psm, "BASiCS_norm")[select,]
  counts_sm <- assay(sce_sm, "BASiCS_norm")[select,]
  rownames(counts_psm) <- rownames(counts_sm) <- 
    rowData(sce_sm)[select, "external_gene_name"]
  list(psm = counts_psm, sm = counts_sm)
}
```

Next, we define a function that uses the `circlize` package to create a
color scale for the heatmaps:

```{r color}
make_colorscale_counts <- function(counts_sm, counts_psm) {
  colorRamp2(
    breaks = seq(0,
      log10(max(c(as.numeric(counts_sm), as.numeric(counts_psm)) + 1)),
      length.out = 20
    ),
    colors = viridis(20)
  )
}
```

Next, we define a function that uses the `circlize` package to create a
color scale for the side annotations displaying the mean expression values in
each group:

```{r mu-color}
make_colorscale_mu <- function(log_mu_sm, log_mu_psm) {
  colorRamp2(
    breaks = seq(0, max(c(log_mu_sm, log_mu_psm)), length.out = 20),
    colors = viridis(20, option = "A", direction = 1)
  )
}
```

Then, we define a function that uses the `ComplexHeatmap` package to
create a single heatmap, with the normalised counts for a single population,
featuring the mean expression values for each gene as a side annotation:

```{r heatmap-drawer}
make_heatmap <- function(counts, col, log_mu, mu_col, title) {
  Heatmap(
    log10(as.matrix(counts) + 1),
    col = col,
    right_annotation = rowAnnotation(
      `log(mu)` = log_mu,
      col = list(`log(mu)` = mu_col)
    ),
    name = "log10(count + 1)",
    column_title = title,
    show_column_names = FALSE,
    cluster_columns = FALSE,
    cluster_rows = FALSE,
    row_names_gp = gpar(fontsize = 6)
  )
}
```


Finally, we define a function that uses the functions we defined so
far to create a combined heatmap showing the normalised counts for both
populations, with mean expression values as side annotations:

```{r heatmap-fun}
plot_heatmap <- function(table, condition, decreasing = TRUE) {
  select <- select_top_n(table_de, condition, decreasing = decreasing)
  counts_sub <- use_select(sce, sce_psm, sce_sm, select)
  counts_sm <- counts_sub$sm
  counts_psm <- counts_sub$psm

  ## Subset table of DE results to extract mean estimates
  match_order <- match(rownames(counts_sm), table_de$external_gene_name)
  table_de_selected <- table_de[match_order, ]

  col <- make_colorscale_counts(counts_sm, counts_psm)
  log_mu_sm <- log10(table_de_selected$Mean1)
  log_mu_psm <- log10(table_de_selected$Mean2)

  mu_col <- make_colorscale_mu(log_mu_sm, log_mu_psm)

  h <- make_heatmap(counts_sm, col, log_mu_sm, mu_col, "Somitic cells") +
      make_heatmap(counts_psm, col, log_mu_psm, mu_col, "Pre-somitic cells")
  draw(h)
}
```

First, we show genes up-regulated in somitic cells:


```{r heatmap-diffexp1, fig.width = 5, fig.height = 3.5, fig.cap="Normalised expression values ($\\log_{10}(x + 1)$ scale) for somitic and pre-somitic cells, showing genes up-regulated in PSM cells. For each group, 10 example genes are shown. These were selected according to the ranking of their associated tail posterior probabilities associated to the differential mean expression test. Colour indicates expression level; colour bars on the right of heatmap segments indicate the inferred mean expression level (in log scale) for each gene in each population."}
plot_heatmap(table_de, "PSM+")
```

We can create a similar plot for genes up-regulated in SM cells:

```{r heatmap-diffexp2, fig.width = 5, fig.height = 3.5, fig.cap="Normalised expression values ($\\log_{10}(x + 1)$ scale) for somitic and pre-somitic cells, showing genes up-regulated in SM cells. For each group, 10 example genes are shown. These were selected according to the ranking of their associated tail posterior probabilities associated to the differential mean expression test. Colour indicates expression level; colour bars on the right of heatmap segments indicate the inferred mean expression level (in log scale) for each gene in each population."}
plot_heatmap(table_de, "SM+")
```


Finally, we show non-differentially expressed genes:

```{r heatmap-diffexp3, fig.width = 5, fig.height = 3.5, fig.cap="Normalised expression values ($\\log_{10}(x + 1)$ scale) for somitic and pre-somitic cells, showing non-differentially expressed genes. For each group, 15 example genes are shown. These were selected according to the ranking of their associated tail posterior probabilities associated to the differential mean expression test. Colour indicates expression level; colour bars on the right of heatmap segments indicate the inferred mean expression level (in log scale) for each gene in each population."}
plot_heatmap(table_de, "NoDiff", decreasing = FALSE)
```

While several computational tools exist to perform differential mean expression
analysis using scRNAseq data [@Soneson2018], the key focus of
`r Biocpkg("BASiCS")` is to perform differential variability testing:
identifying changes in transcriptional variability between the groups of cells. 
To avoid the confounding between mean and over-dispersion, we recommend 
to use residual over-dispersion parameters $\epsilon_i$ as input to this 
analysis. 

We can now visualise the changes in residual over-dispersion between somitic and 
pre-somitic mesoderm cells in the form of a MA-plot 
(Figure \@ref(fig:diff-res-plot)).
In this visualisation, the difference between the posterior medians of the 
residual over-dispersion parameters $\epsilon_i$ are shown on the y-axis.
Epsilon values for genes that are not expressed in at least 2 cells per 
conditions are marked as `NA` and are therefore not displayed.

```{r diff-res-plot, fig.height = 8, fig.width = 6, fig.cap = "Upper panel presents the MA plot associated to the differential residual over-dispersion test between somitic and pre-somitic cells. Differences of residual over-dispersion in somitic cells relative to pre-somitic mesoderm cells are plotted against average expression estimates combined across both groups of cells. Bottom panel presents the volcano plot associated to the same test. Differences of residual over-dispersion in somitic cells relative to pre-somitic cells are plotted against their associated tail posterior probabilities. Colour indicates the differential expression status for each gene, including a label to identify genes that were excluded from differential expression test due to low ESS."}
p1 <- BASiCS_PlotDE(test_de, Parameters = "ResDisp", Plots = "MA")
p2 <- BASiCS_PlotDE(test_de, Parameters = "ResDisp", Plots = "Volcano",
  TransLogit = TRUE)
p1 / p2
```

While one could focus on the sets of gene that show significant changes in
residual over-dispersion, here we want to highlight how to analyse
changes in mean expression in parallel to changes in variability.
For this, we first combine the results of the differential mean expression 
and the differential residual over-dispersion test.
These are independent analyses, given that changes in residual
over-dispersion are not confounded with changes
in mean expression, as shown in Figure \@ref(fig:combine-results).
<!---We further remove the genes that were excluded from the test and those 
that are not expressed in at least 2 cells in either condition.---> 

```{r combine-results, fig.cap="Difference in residual over-dispersion against $\\log_2$ fold change in mean expression."}
## create list of generic plot parameters to use across a few plots
plot_params <- list(
  geom_pointdensity(na.rm = TRUE),
  scale_colour_viridis(name = "Density"),
  theme(
    # text = element_text(size = rel(3)),
    legend.position = "bottom",
    legend.text = element_text(angle = 45, size = 8, hjust = 1, vjust = 1),
    legend.key.size =  unit(0.018, "npc")
  )
)
## plot log2FC against difference of residual over-dispersion
ggplot(table_de) +
  aes(MeanLog2FC, ResDispDistance) +
  plot_params +
  xlim(-15, 15) +
  labs(
    x = "log2 fold change in mean expression",
    y = "Difference in residual over-dispersion"
  )
```

While genes with significant changes in residual over-dispersion often have
similar levels of mean expression, as seen in
Figure \@ref(fig:var-plots)A and C,
they may have a different proportion of zero counts in the two cell populations.
Figure \@ref(fig:var-plots)B and D show that many genes with higher residual
over-dispersion in somitic cells have a lower proportion of zeros in pre-somitic
mesoderm cells,
and vice versa.

```{r, var-plots, fig.width = 6, fig.height = 4, fig.cap="A, C: $\\log_2$ change in expression against log mean expression for genes with higher residual over-dispersion in somitic (A) cells and pre-somitic mesoderm (C) cells. B, D: Proportion of expressed cells for genes, with higher residual over-dispersion in somitic cells (B) and pre-somitic mesoderm (D) cells. Dashed red lines in panels A and C represent a log fold change of zero, meaning no change in average expression. Dashed red lines in panels B and D represent the line described by y=x, representing equal detection levels in both populations."}
## mean expression vs log2FC for genes with higher residual over-dispersion in
## somitic cells
g1 <- ggplot(
    table_de[table_de$ResultDiffResDisp == "SM+", ]
  ) +
  aes(MeanOverall, MeanLog2FC) +
  plot_params +
  ylim(-15, 15) +
  scale_x_log10() +
  labs(x = "Mean expression", y = "log2 fold change") +
  geom_hline(
    yintercept = 0, colour = "firebrick", linetype = "dashed"
  )

## mean expression vs log2FC for genes with higher residual over-dispersion in
## pre-somitic cells
g2 <- ggplot(
    table_de[table_de$ResultDiffResDisp == "PSM+", ]
  ) +
  aes(MeanOverall, MeanLog2FC) +
  plot_params +
  ylim(-15, 15) +
  scale_x_log10() +
  labs(x = "Mean expression", y = "log2 fold change") +
  geom_hline(
    yintercept = 0, colour = "firebrick", linetype = "dashed"
  )

g1 + g2 + plot_annotation(tag_levels = "A")
```

Similarly to analysis of differential expression, it is useful to visualise 
the results of differential variability tests
in order to appraise the quality
of the results, and to identify systematic patterns among the genes identified. 
One useful way to do this is by examining the normalised counts on a
gene-by-gene basis. To do this, we first add some gene-level summary statistic
about the normalised counts to the `table_de` object,
and then define a function that
selects the normalised counts for a given gene, and then reshapes them into
a form suitable for plotting. 


<!---These are genes with higher variability in one population
than the other, but which have a similar frequency of detection in each
population. Given that these genes have a similar proportion of zero counts
in both populations, it is likely that the differences in variability
are not driven by a more intermittent gene expression pattern.

These are genes with higher variability in one population and differing levels of detection; as mentioned above,
these may represent genes with a more intermittent expression
pattern in one cell population. These may point to
differences in intrinsic sources of variability in expression, such as regulatory
elements (e.g., transcription factors). They may also represent genes that
are markers of extrinsic variability, for example cell sub-populations that 
differ in abundance between the cell populations in question. Disentangling
intrinsic and extrinsic sources of variability when interpreting such genes is 
an important component of any analysis of differential variability, and should
be undertaken with care.
--->


```{r vg-df-fun}
## merge with some summary statistics about genes
gene_tests <- data.frame(
  "GeneName" = rownames(sce_sm),
  "ExpPropSM" = rowMeans(assay(sce_sm, "BASiCS_norm") > 0),
  "ExpPropPSM" = rowMeans(assay(sce_psm, "BASiCS_norm") > 0)
)

table_de_genetests <- merge(table_de, gene_tests)

make_vg_df <- function(ind_vg) {
  table_de_vg <- table_de[ind_vg, ]
  ## If more than 4 genes,
  ## pick top 4 ranked by differences in residual over-dispersion
  if(nrow(table_de_vg) > 4) {
    table_de_vg <- table_de_vg[
      order(abs(table_de_vg$ResDispDistance), decreasing = TRUE), ]
    table_de_vg <- table_de_vg[1:4, ]
  }
  var_genes <- table_de_vg$GeneName
  var_sm <- as.matrix(assay(sce_sm, "BASiCS_norm")[var_genes, , drop = FALSE])
  var_psm <- as.matrix(assay(sce_psm, "BASiCS_norm")[var_genes, , drop = FALSE])
  var_df <- rbind(
    data.frame(t(var_sm), Group = "SM"),
    data.frame(t(var_psm), Group = "PSM")
  )
  var_gene_names <- rowData(sce_sm)[var_genes, "external_gene_name"]
  colnames(var_df)[-ncol(var_df)] <- var_gene_names
  pivot_longer(var_df, cols = var_gene_names)
}
```

We can use this function to wrangle the data for a selection of genes we
want to visualise. For example, all genes that are identified as differentially
variable in somitic cells:

```{r vg-df}
df_vg_sm <- make_vg_df(table_de_genetests$ResultDiffResDisp %in% c("SM+"))
head(df_vg_sm)
```

We can then use this function to plot the normalised counts for genes
that fill a number of criteria with respect to mean expression, expression
variability, and detection levels, by defining a function that
uses the `make_vg_df` function we just defined. This function takes
a logical vector and uses it to subset the `table_de`, `sce_sm` and `sce_psm`
objects, merges them into a form suitable for plotting, and then plots them.

```{r vg-plot-fun}
## Utility function that plots logcounts of a set of genes defined by "ind_vg"
plot_vg <- function(ind_vg) {
  df <- make_vg_df(ind_vg)
  ggplot(df, aes(x = name, y = value + 1, colour = Group)) +
    geom_violin(width = 0.8, position = position_dodge(width = 0.8)) +
    geom_point(
      position = position_jitterdodge(jitter.width = 0.2, dodge.width = 0.8),
      alpha = 0.2
    ) +
    scale_x_discrete(guide = guide_axis(angle = 45)) +
    scale_y_log10() +
    labs(x = "Gene", y = "Denoised count + 1") +
    scale_colour_brewer(palette = "Set1")
}
```

First, we select genes that are more variable in somitic cells than in
pre-somitic mesoderm cells, but which are not differentially expressed.
Furthermore, we narrow our selection to genes with
similar levels of non-zero expression in both populations, and which are
expressed in at least 50% of cells within each population. These criteria
are designed to select genes that likely have a largely unimodal expression
pattern, simply having higher cell-to-cell variability in somitic cells.
We then plot the normalised counts for
these genes using the function we defined above, and save it to
a variable called `g1` for display later.


```{r select-vg1}
ind_vg_unimodal_sm <- table_de_genetests$ResultDiffResDisp %in% c("SM+") &
  table_de_genetests$ResultDiffMean %in% c("NoDiff") &
  abs(
    table_de_genetests$ExpPropSM - table_de_genetests$ExpPropPSM
  ) < 0.05 &
  pmin(
    table_de_genetests$ExpPropSM, table_de_genetests$ExpPropPSM
  ) > 0.99
g1 <- plot_vg(ind_vg_unimodal_sm) +
    theme(legend.position = "none")
```

Next, we select genes that are more variable in somitic cells
than in pre-somitic mesoderm cells, but which are not differentially expressed.
Furthermore, we narrow our selection to genes with
different levels of non-zero expression in both populations, and which are
expressed in at least 50% of cells within one or more of the populations.
These are genes that could be described as having a more bursting pattern
of expression in somitic cells. We then plot the normalised counts for
these genes, and again save the plot object for display later.

```{r select-vg2}
ind_vg_bursting_sm <- table_de_genetests$ResultDiffResDisp %in% c("SM+") &
  table_de_genetests$ResultDiffMean %in% c("NoDiff") &
  abs(
    table_de_genetests$ExpPropSM - table_de_genetests$ExpPropPSM
  ) > 0.1 &
  pmax(
    table_de_genetests$ExpPropSM, table_de_genetests$ExpPropPSM
  ) > 0.75
g2 <- plot_vg(ind_vg_bursting_sm)
```

Now, we use the same selection criteria, plotting first for genes
that are largely unimodal, with a high cell-to-cell variability in
pre-somitic cells, and then for genes that have a more bursting
expression pattern in pre-somitic cells:

```{r select-vg3}
ind_vg_unimodal_psm <- table_de_genetests$ResultDiffResDisp %in% c("PSM+") &
  table_de_genetests$ResultDiffMean %in% c("NoDiff") &
  abs(
    table_de_genetests$ExpPropSM - table_de_genetests$ExpPropPSM
  ) < 0.1 &
  pmin(
    table_de_genetests$ExpPropSM, table_de_genetests$ExpPropPSM
  ) > 0.75

g3 <- plot_vg(ind_vg_unimodal_psm) +
    theme(legend.position = "none")


ind_vg_bursting_psm <- table_de_genetests$ResultDiffResDisp %in% c("PSM+") &
  table_de_genetests$ResultDiffMean %in% c("NoDiff") &
  abs(
    table_de_genetests$ExpPropSM - table_de_genetests$ExpPropPSM
  ) > 0.10 &
  pmax(
    table_de_genetests$ExpPropSM, table_de_genetests$ExpPropPSM
  ) > 0.50
g4 <- plot_vg(ind_vg_bursting_psm)
```

Now that we have created four violin plots of the various types of
differentially variable gene that we have defined, we can combine the plots
together to visualise them simultaneously:


```{r violin-diffresdisp, fig.width = 7, fig.height = 5, fig.cap="Violin plots of denoised counts. A: Four genes with higher residual over-dispersion in somitic cells, and similar levels of detection in pre-somitic and somitic populations. B: Four genes with higher residual over-dispersion in somitic cells and different levels of detection in somitic and pre-somitic cells. C: Four genes with higher residual over-dispersion in pre-somitic cells and similar levels of detection in somitic and pre-somitic cells. D: Three genes with higher residual over-dispersion in pre-somitic cells and different levels of detection in somitic and pre-somitic cells."}
(g1 + g2) / (g3 + g4) + plot_annotation(tag_levels = "A")
```

Figure \@ref(fig:violin-diffresdisp) shows denoised counts for
genes with significant differences in residual over-dispersion, with each panel 
showing a different type of expression patterns. 
Exploration of such patterns is important component of any analysis of 
differential variability, and should be undertaken with care.
Figure \@ref(fig:violin-diffresdisp)B and \@ref(fig:violin-diffresdisp)D
show genes with differing levels of detection in both populations, as well as 
higher levels of residual over-dispersion in somitic or pre-somitic mesoderm cells (B and D, 
respectively).
Thus, these genes may represent those with a more bursty expression pattern in 
one of the cell population. 
They may also represent genes that are markers of extrinsic variability, 
for example cell sub-populations that differ in abundance between the cell populations in question. 
In contrast, Figure
\@ref(fig:violin-diffresdisp)A and \@ref(fig:violin-diffresdisp)C
show genes with similar levels of detection in both populations, as well as 
higher levels of residual over-dispersion in somitic or pre-somitic mesoderm cells (A and C, 
respectively). 
These cases are likely driven by more tight regulation, rather than 
transcriptional burst or sub-population structure. 




# Discussion 

In this article, we have explored the research questions that 
`r Biocpkg("BASiCS")` seeks to 
resolve --- chiefly, robustly quantifying average and variability in
expression in cell populations. We have outlined the appropriate quality control
and data visualisation steps to apply when undertaking an analysis using 
`r Biocpkg("BASiCS")`
in order to ensure high quality input data.
We have also outlined the steps needed to use `r Biocpkg("BASiCS")`
to quantify biological variability, identify highly variable genes, and
normalise scRNAseq data from a single population. We have also provided a 
limited comparison of the results of these analyses using `r Biocpkg("BASiCS")`
and the result of similar analyses using `r Biocpkg("scran")`.
Furthermore, we have demonstrated functions within `r Biocpkg("BASiCS")`
that allow users to ensure the MCMC used in `r Biocpkg("BASiCS")`
has converged and produced adequate sample sizes. Finally, we have 
demonstrated the use of `r Biocpkg("BASiCS")` to robustly identify 
differentially expressed genes, in terms of mean expression and in terms of 
biological variability.

Further challenges exist in analysing scRNAseq data [@Eling2019;@Lahnemann2020]. 
One area for future work is to study how post-selection inference may affect the 
performance of `r Biocpkg("BASiCS")`. For example, @Neufeld2022 have developed
a framework for addressing inflated type I errors when studying differences in
means between groups of cells identified via hierarchical clustering,
by splitting counts into "train" and "test" sets.
Another important challenge 
for `r Biocpkg("BASiCS")` is computational efficiency. The number of cells 
profiled in scRNAseq experiments has scaled exponentially since the development 
of the technology [@Svensson2018].
Given that `r Biocpkg("BASiCS")` requires computationally intensive
MCMC sampling to estimate the posterior distribution, it becomes computationally
intractable to analyse data from very large numbers of cells. 
Alternative inference schemes such as variational inference, maximum
*a posteriori* estimation, or Hamiltonian Monte Carlo may be useful in this
context. Advanced users may wish to use the `r Biocpkg("BASiCStan")` package
to test these alternative inference schemes. This also provides access to
the model diagnostics, facilities for running multiple chains, and
posterior summaries provided by Stan [@Carpenter2017], while also being
fully compatible with the workflow described in this manuscript via the function
`Stan2BASiCS`, that converts the output of the Stan inference procedure to
the type of output generated by `BASiCS_MCMC`. However, we note that
the inference methods provided by Stan are often very slow (Hamiltonian
Monte Carlo) or unreliable (variational inference, maximum *a posteriori*)
for BASiCS.

Finally, we also anticipate potential extensions of `r Biocpkg("BASiCS")` to 
account for the more complex experimental designs. For example, in cohort 
studies where cells are extracted from multiple individuals, the hierarchical 
model could be expanded to quantify both for intra- and inter-individual 
transcriptional variability. As spatial transcriptomic technologies mature 
[@Marx2021], Bayesian hierarchical models such as the one implemented in 
`r Biocpkg("BASiCS")` could also be designed to incorporate spatial (and, 
potentially, temporal) structure (e.g. [@Aijo2019]). 
We intend to update this workflow as the field evolves, and as we address the 
issues and challenges outlined here. 

# Reproducibility

The following software versions were used throughout this workflow:

- **R version**: `r R.version.string`

- **Bioconductor version**: `r BiocManager::version()`

- **R packages**: 
  - BASiCS `r packageVersion("BASiCS")`
  - scran `r packageVersion("scran")`
  - scater `r packageVersion("scater")`  
  Version numbers for all remaining packages are available in the
  [Session Info](#session-info) section.

A Docker image containing all software requirements is available at 
[Docker hub](https://hub.docker.com/repository/docker/alanocallaghan/bocker).

# Session Info {#session-info}

```{r, session-info}
sessionInfo()
```


# Data availability {#data-availability}

The data used in this article are available for download
on EBI ArrayExpress under accession number
[E-MTAB-4888](https://www.ebi.ac.uk/arrayexpress/experiments/E-MTAB-4888/).
The MCMC chains used to generate this article can be found in Zenodo
under the DOI [10.5281/zenodo.10251224](https://doi.org/10.5281/zenodo.10251224).

# Software availability

**Software: All software used in this workflow is available as part of 
Bioconductor 3.18.**

This section will be generated by the Editorial Office before publication. 
Authors are asked to provide some initial information to assist the 
Editorial Office, as detailed below.

1. URL link to where the software can be downloaded from or used by a non-coder
  https://bioconductor.org/packages/3.13/bioc/html/BASiCS.html
  (AUTHOR TO PROVIDE; optional)
2. URL link to the author's version control system repository containing the 
  source code (AUTHOR TO PROVIDE; required)
  Software: https://github.com/catavallejos/BASiCS/
  Manuscript: https://github.com/VallejosGroup/BASiCSWorkflow/
3. Link to source code as at time of publication (*F1000Research* TO GENERATE)
4. Link to archived source code as at time of publication (*F1000Research* 
  TO GENERATE)
5. Software license (AUTHOR TO PROVIDE; required)
  GPL (>= 2)

# Author information

N.E. and A.OC. made the primary contributions to the manuscript.
C.A.V. and J.C.M. provided guidance and edited the manuscript.

# Competing interests

No competing interests were disclosed.

# Grant information

C.A.V. is a Chancellor's Fellow funded by the University of Edinburgh.
A.OC. was funded by the Chancellor's Fellowship granted to C.A.V.
N.E. was funded by the European Molecular Biology Laboratory International PhD
Programme. Work by JCM was supported by CRUK (C9545/A29580) and by core support
from EMBL.
